---
title: "Logistic Regression for Loansharks (2015 IL)"
author: "Anantanarayanan G Iyengar"
date: "4/27/2020"
output: html_notebook
---

## We only tried one glm model here. This was based on the work done
## for the models for Pennsylvania. 
# Model 1. Log of loan amount and loan to income ratio.

## Results.
# Performance was similar to the models for PA. The criterion was AUC 
# curve, Mcfadden R square value and confusion matrix results on test data.

## Key observation
# The AUC curve for our model is around 74.5. The AUC curve can
# be thought of as a classification-threshold-invariant. It measures the #
# quality of the model's predictions irrespective of the threshold value
# https://developers.google.com/machine-learning/crash-course/classification/roc-and-auc.
# We feel that this is because our models miss crucial data like debt to income
# ratio, credit scores, etc which banks use as additional predictors.


## Global setup like working directory, data directory etc should happen here.

```{r}
library(sys)

working_directory <- getwd()

setwd(dirname(dirname(working_directory)))

writeLines("")
getwd()
# The HMDA dataset is loaded from here.
data_dir <- "D:/data"
census_data_dir = "d:/data/census"

# The models are saved in this directory.
model_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "models", sep = "/")
print(model_path)

# The train and test data are saved in this directory
model_data_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "data", sep = "/")
print(model_data_path)

options(scipen = 999)

```

## Install required packages.
```{r}
# https://stackoverflow.com/questions/4090169/elegant-way-to-check-for-missing-packages-and-install-them
list_of_packages <- c("mlbench", "corrplot", "rvest", "tidyr", "stringr", "dplyr", "lubridate", "data.table", "mice", "scales", "naniar", "rpart", "rpart.plot", "caret", "pROC", "ROCR")
new.packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]

if (length(new.packages)) {
  print("Installing packages\n")
  install.packages(new.packages())
}

library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)
library(caret)
library(pROC)
library(ROCR)

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/utils.r", sep="/"))

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/model_utils.r", sep="/"))

```

## Load data file
```{r}
#hmda_data_pa <- fread(paste(data_dir, "/hmda_2007_pa_all-records_labels.csv", sep = ""))

hmda_data_il <- fread(paste(data_dir, "/HMDA/2015/hmda_2015_il_all-records_labels.csv", sep = ""))
```

## Convert to dataframe
```{r}
hmda_il_df <- as.data.frame(hmda_data_il)
head(hmda_il_df)
```


# Impute as needed.
```{r}

# https://www.rdocumentation.org/packages/mice/versions/3.8.0/topics/mice.impute.cart
hmda_il_df_imputed <- mice(hmda_il_df, m=1, maxit=2, meth='cart',seed=500)

hmda_data_il_df_imputed <- mice::complete(hmda_il_df_imputed)

```

```{r}
## Add a new column applicant_race_and_ethnicity and group all applicants
## with ethnicity as Hispanic or Latino as Hispanic or Latino in this column.
## For everyone else, this column gets values from the applicant_race_name_1 
## column

hmda_data_il_df_imputed <- add_applicant_race_and_ethicity_to_hmda(hmda_data_il_df_imputed)
```


# Save the imputed dataframe to file.
```{r}

write.csv(hmda_data_il_df_imputed, paste(data_dir, "/HMDA/2015/hmda_2015_il_imputed.csv", sep = ""), row.names = FALSE)
```

# Explore the dataframe obtained above.
```{r}
hmda_data_df <- hmda_data_il_df_imputed
head(hmda_data_df, 20)

```

# Count how many unique county codes and respondent ids there are
```{r}

length(unique(hmda_data_df$county_code))
length(unique(hmda_data_df$respondent_id))
```

# Create dataframe based on selected column list from the source dataframe.
```{r}
hmda_data_df_for_model <- hmda_data_frame_for_model(hmda_data_df)
hmda_data_df_for_model <- process_model_df_columns(hmda_data_df_for_model)

```

# Explore the dataframe with selected columns obtained above.

```{r}
head(hmda_data_df_for_model, 20)

nrow(hmda_data_df_for_model)

length(levels(hmda_data_df_for_model$respondent_id))
length(levels(hmda_data_df_for_model$county_code))


```

# Train/test split.
```{r}
set.seed(400)

split_dfs <- train_test_split(hmda_data_df_for_model$loan_granted, hmda_data_df_for_model)

train <- split_dfs[[1]]
test <- split_dfs[[2]]

```


# Null model
```{r}
null_model <- build_and_evaluate_hmda_null_model("loan_granted~1", train, test)
```


# Model with logs of applicant income and loan amount and loan to income ratio

```{r}
model_2015_il_ratio_of_loan_to_income_with_log_of_income_and_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+log_loan_amount_000s+applicant_sex+county_code+tract_to_msamd_income+co_applicant_present+agency_code+minority_population+loan_purpose+rate_spread+property_type++loan_to_income_ratio+respondent_id", train)

model_summary_and_plot(model_2015_il_ratio_of_loan_to_income_with_log_of_income_and_amount, train)

mcfadden_r_square(model_2015_il_ratio_of_loan_to_income_with_log_of_income_and_amount, null_model)

```

# Model performance.

```{r}
library(ROCR)
library(pROC)
model1_fitted.probs <- predict(model_2015_il_ratio_of_loan_to_income_with_log_of_income_and_amount, test, type = "response")

writeLines("")
writeLines("Model 1 performance")
print_model_performance(model1_fitted.probs, test, positive = "yes", negative = "no", threshold = 0.6)

```

# Save the models
```{r}
dir.create(model_path, showWarnings = FALSE)
save(model_2015_il_ratio_of_loan_to_income_with_log_of_income_and_amount, file = paste(model_path, "/model_il_loan_income_ratio_with_log_of_income_and_amount.rdata", sep=""))
```

# Save the train and test data for future reference.
```{r}

dir.create(model_data_path, showWarnings = FALSE)
save(train, file = paste(model_data_path, "/train_il.rdata", sep = ""))
save(test, file = paste(model_data_path, "/test_il.rdata", sep = ""))

```
