---
title: "PA_2015_Random_Forest"
author: "Virat Joshi"
date: "4/24/2020"
output: html_document
---


# Set the working and data directories
```{r}
working_dir = "/Users/vjstark/Downloads/CSP-571/Project"
setwd(working_dir)

data_dir <- "./Data"
```

```{r}
# The models are saved in this directory.
model_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "models", sep = "/")
print(model_path)

# The train and test data are saved in this directory
model_data_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "data", sep = "/")
print(model_data_path)

```

# Load Essential Libraries
```{r}
library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)
library(caret)
library(pROC)
library(ROCR)
library(precrec)

utils_path = './utils/utils.r'
model_utils_path = './utils/model_utils.r'
source(paste(utils_path))
source(paste(model_utils_path))
```


# Load the imputed hmda csv file.
```{r}
hmda_data_df_ <- fread(paste(data_dir, "/hmda_2015_pa_imputed.csv", sep=""))
hmda_data_df <- as.data.frame(hmda_data_df_)
```

# Explore the dataframe obtained above.
```{r}
head(hmda_data_df, 20)
colnames(hmda_data_df)
```



# Create dataframe based on selected column list from the source dataframe.
```{r}
hmda_data_df_for_model <- hmda_data_frame_for_model(hmda_data_df)
hmda_data_df_for_model <- process_model_df_columns(hmda_data_df_for_model)

```


# Explore the dataframe with selected columns obtained above.
```{r}
head(hmda_data_df_for_model, 20)

nrow(hmda_data_df_for_model)

```

# Train/test split.
```{r}
set.seed(400)

split_dfs <- train_test_split(hmda_data_df_for_model$loan_granted, hmda_data_df_for_model)

train <- split_dfs[[1]]
test <- split_dfs[[2]]
```

From the selected columns in EDA for feature importances, we select top features and build a model using those predictors.
```{r}
library("randomForest")
set.seed(1234)

rf_formula = loan_granted~applicant_race_and_ethnicity + applicant_income_000s + loan_amount_000s + preapproval + co_applicant_present + minority_population + loan_purpose + tract_to_msamd_income + applicant_sex + property_type + owner_occupancy + agency_code + rate_spread + respondent_id

rf_model <- randomForest(rf_formula,data=train,ntree=128)
```

```{r}
pred = predict(rf_model, newdata=test)
#table(pred, test$loan_granted)
confusionMatrix(pred,test$loan_granted, positive = "yes")
precision_recall(confusionMatrix(pred,test$loan_granted, positive = "yes")$table)
```

# A plot of error vs the number of tree.
```{r}
plot(rf_model)
```

# Importance of each parameter in terms of their Gini values
```{r}
importance(rf_model)
```