---
title: "Logistic Regression for Loansharks (2014 Pennsylvania)"
author: "Anantanarayanan G Iyengar [Please add your names here]" 
date: "4/5/2020"
output: html_notebook
---

## We tried the following glm models. The predictors chosen were picked
## from our EDA analysis and feature engineering work and some background
## knowledge.
# Model 1. Raw loan and applicant income.
# Model 2. Log of loan and applicant income.
# Model 3. Log of loan amount and loan to income ratio.
# Model 4. Cross validation model with same predictors as Model 3.
# Model 5. Merged model with hmda data and census demographic and ethnicity
#          data. We removed the minority_population column from the HMDA data
#          and replaced it with predictors like BLACK_PERCENT, ASIAN_PERCENT, #            etc. We were trying to see if African Americans were negatively 
#          impacted if they applied for loans in areas where the BLACK_PERCENT
#          was high, etc.
# Model 6. Merged model with median loan to value ratio coming from Zillow.
#          Banks use the loan to value ratio as one of the criterion in loan
#          decisions. As HMDA data does not have this information, we thought
#          we could approximate this by using the median home values in the
#          county. 

## Results.
# Model 2, Model 3 and Model 4 performed similarly. The criterion was AUC 
# curve, Mcfadden R square value and confusion matrix results on test data.
# Model 5 and Model 6 also performed well. However the results weren't that 
# different from the simpler models. Hence we opted for the simpler models.

# Model 3 was chosen as banks do use the loan to income ratio as one of their
# criterion. 

## Key observation
# The AUC curve for our models ranges between 75 to 77%. The AUC curve can
# be thought of as a classification-threshold-invariant. It measures the #
# quality of the model's predictions irrespective of the threshold value
# https://developers.google.com/machine-learning/crash-course/classification/roc-and-auc.
# We feel that this is because our models miss crucial data like debt to income
# ratio, credit scores, etc which banks use as additional predictors.


## Global setup like working directory, data directory etc should happen here.

```{r}
library(sys)

working_directory <- getwd()

setwd(dirname(dirname(working_directory)))

writeLines("")
getwd()
# The HMDA dataset is loaded from here.
data_dir <- "D:/data"
census_data_dir = "d:/data/census"

# The models are saved in this directory.
model_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "models", sep = "/")
print(model_path)

# The train and test data are saved in this directory
model_data_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "data", sep = "/")
print(model_data_path)

options(scipen = 999)

```

## Install required packages.
```{r}
# https://stackoverflow.com/questions/4090169/elegant-way-to-check-for-missing-packages-and-install-them
list_of_packages <- c("mlbench", "corrplot", "rvest", "tidyr", "stringr", "dplyr", "lubridate", "data.table", "mice", "scales", "naniar", "rpart", "rpart.plot", "caret", "pROC", "ROCR")
new.packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]

if (length(new.packages)) {
  print("Installing packages\n")
  install.packages(new.packages())
}

library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)
library(caret)
library(pROC)
library(ROCR)

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/utils.r", sep="/"))

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/model_utils.r", sep="/"))

```

# Load the imputed hmda csv file.
```{r}
hmda_data_df_ <- fread(paste(data_dir, "/HMDA/2014/hmda_2014_pa_imputed.csv", sep=""))
hmda_data_df <- as.data.frame(hmda_data_df_)
```

# Explore the dataframe obtained above.
```{r}
head(hmda_data_df, 20)

```

# Create dataframe based on selected column list from the source dataframe.
```{r}
hmda_data_df_for_model <- hmda_data_frame_for_model(hmda_data_df)
hmda_data_df_for_model <- process_model_df_columns(hmda_data_df_for_model)

```

# Explore the dataframe with selected columns obtained above.

```{r}
head(hmda_data_df_for_model, 20)

nrow(hmda_data_df_for_model)

length(levels(hmda_data_df_for_model$respondent_id))
length(levels(hmda_data_df_for_model$county_code))


```

# Train/test split.
```{r}
set.seed(400)

split_dfs <- train_test_split(hmda_data_df_for_model$loan_granted, hmda_data_df_for_model)

train <- split_dfs[[1]]
test <- split_dfs[[2]]

```


# Null model
```{r}
null_model <- build_and_evaluate_hmda_null_model("loan_granted~1", train, test)
```

# Model with applicant income and loan amount.
```{r}
model_2014_income_loan_amount <-
build_glm_model("loan_granted~applicant_race_and_ethnicity+applicant_income_000s+loan_amount_000s+preapproval+co_applicant_present+county_code+minority_population+loan_purpose+tract_to_msamd_income+applicant_sex+property_type+owner_occupancy+agency_code+rate_spread+respondent_id", train)

model_summary_and_plot(model_2014_income_loan_amount, train)
mcfadden_r_square(model_2014_income_loan_amount, null_model)

```


# Model with logs of applicant income and loan amount.
```{r}
model_2014_log_income_loan_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+log_loan_amount_000s+preapproval+co_applicant_present+county_code+tract_to_msamd_income+loan_purpose+owner_occupancy+respondent_id+minority_population+rate_spread+agency_code+applicant_sex+property_type", train)

model_summary_and_plot(model_2014_log_income_loan_amount, train)
mcfadden_r_square(model_2014_log_income_loan_amount, null_model)

```

# Model with logs of applicant income and loan amount and loan to income ratio

```{r}
model_2014_ratio_of_loan_to_income_with_log_of_income_and_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+log_loan_amount_000s+applicant_sex+county_code+tract_to_msamd_income+co_applicant_present+agency_code+minority_population+loan_purpose+rate_spread+property_type++loan_to_income_ratio+respondent_id", train)

model_summary_and_plot(model_2014_ratio_of_loan_to_income_with_log_of_income_and_amount, train)

mcfadden_r_square(model_2014_ratio_of_loan_to_income_with_log_of_income_and_amount, null_model)

```


# Cross validation
```{r}
model_2014_cross_val_loan_income_ratio <- cross_validation("loan_granted~applicant_race_and_ethnicity+loan_to_income_ratio+co_applicant_present+owner_occupancy+preapproval+property_type+respondent_id+log_loan_amount_000s+county_code+loan_purpose+tract_to_msamd_income+minority_population+agency_code+rate_spread+loan_to_income_ratio", train)

model_summary_and_plot(model_2014_cross_val_loan_income_ratio$finalModel, train)

```

# Model performance.

```{r}
library(ROCR)
library(pROC)
model1_fitted.probs <- predict(model_2014_income_loan_amount, test, type = "response")
model2_fitted.probs <- predict(model_2014_log_income_loan_amount, test, type = "response")
model3_fitted.probs <- predict(model_2014_ratio_of_loan_to_income_with_log_of_income_and_amount, test, type = "response")

writeLines("")
writeLines("Model 1 performance")
print_model_performance(model1_fitted.probs, test, positive = "yes", negative = "no", threshold = 0.6)

writeLines("")
writeLines("Model 2 performance")
print_model_performance(model2_fitted.probs, test, positive = "yes", negative = "no", threshold = 0.6)

writeLines("")
writeLines("Model 3 performance")
print_model_performance(model3_fitted.probs, test, positive = "yes", negative = "no", threshold = 0.6)

```


# Cross validation performance
```{r}
library(pROC)

mcfadden_r_square(model_2014_cross_val_loan_income_ratio$finalModel, null_model)

model4_fitted.probs <- predict(model_2014_cross_val_loan_income_ratio, test, type = "prob")
writeLines("")
writeLines("Cross validation model performance")

writeLines("")
writeLines("Model 3 performance")
print_model_performance(model4_fitted.probs[, 2], test, positive = "yes", negative = "no", threshold = 0.6)
```

# Model with census data integrated.
```{r}
# https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/counties/asrh/
census_pa_file_path <- paste(census_data_dir, sep = "", "/cc-est2018-alldata-42.csv")

merged_hmda_data_df_for_model <- merge_hmda_df_with_census(census_pa_file_path, hmda_data_df_for_model)

split_dfs <- train_test_split(merged_hmda_data_df_for_model$loan_granted, merged_hmda_data_df_for_model)

merged_train <- split_dfs[[1]]
merged_test <- split_dfs[[2]]

merged_2014_model <- build_glm_model("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+property_type+log_loan_amount_000s+tract_to_msamd_income+co_applicant_present+agency_code+BLACK_PERCENT+ASIAN_PERCENT+HISPANIC_PERCENT+county_code+respondent_id+applicant_sex+loan_purpose+loan_to_income_ratio", merged_train)

model5_fitted.probs <- predict(merged_2014_model, merged_test, type = "response")

writeLines("")
writeLines("Merged model performance")

print_model_performance(model5_fitted.probs, merged_test, positive = "yes", negative = "no", threshold = 0.6)
```

# Loan to value ratio is an important metric which banks use to decide
# whether a loan should be granted or not. For e.g. the higher the loan
# to value ratio the riskier it is for the bank.
# https://en.wikipedia.org/wiki/Loan-to-value_ratio
# Banks are not required  to report this metric in the HMDA reporting
# law. We use the median of the home values in that county based on
# data available from zillow https://www.zillow.com/research/data/
```{r}
period_list <- c("2014-01", "2014-02", "2014-03", "2014-04",
                 "2014-05", "2014-06", "2014-07", "2014-08",
                 "2014-09", "2014-10", "2014-11", "2014-12")

merged_hmda_with_loan_to_value_ratio <- augment_hmda_data_with_estimated_loan_to_value_ratio_from_zillow(paste(data_dir, "/zillow/counties_home_value_per_year.csv", sep = ""), hmda_data_df_for_model, "PA", period_list)


head(merged_hmda_with_loan_to_value_ratio)

```

# Train test split on the hmda dataset merged with estimated loan to value
# ratio
```{r}
split_dfs <- train_test_split(merged_hmda_with_loan_to_value_ratio$loan_granted, merged_hmda_with_loan_to_value_ratio)
new_train <- split_dfs[[1]]
new_test <- split_dfs[[2]]
```

## Model with estimated loan to value ratio.
```{r}
model_2014_with_loan_to_value_ratio <-
build_glm_model("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+log_loan_amount_000s+applicant_sex+county_code+tract_to_msamd_income+co_applicant_present+agency_code+minority_population+loan_to_income_ratio+loan_purpose+rate_spread+property_type+loan_to_value_ratio+respondent_id", new_train)

model_summary_and_plot(model_2014_with_loan_to_value_ratio, train = new_train)
mcfadden_r_square(model_2014_with_loan_to_value_ratio, null_model)

ggplot(data = merged_hmda_with_loan_to_value_ratio, aes(x=loan_to_value_ratio)) +geom_histogram(fill = "steelblue") + labs(title = "Loan amount to Home Value Ratio" , x = "Loan amount to Home Value Ratio",y="Count")

model6_fitted.probs <- predict(model_2014_with_loan_to_value_ratio, new_test, type = "response")

writeLines("")
writeLines("Model with estimated loan to value ratio performance")

print_model_performance(model6_fitted.probs, new_test, positive = "yes", negative = "no", threshold = 0.6)

```

# Save the models
```{r}
dir.create(model_path, showWarnings = FALSE)
save(model_2014_income_loan_amount, file = paste(model_path, "/model_income_loan_amount.rdata", sep=""))
save(model_2014_log_income_loan_amount, file = paste(model_path, "/model_log_income_loan_amount.rdata", sep=""))
save(model_2014_ratio_of_loan_to_income_with_log_of_income_and_amount, file = paste(model_path, "/model_loan_income_ratio_with_log_of_income_and_amount.rdata", sep=""))
save(model_2014_cross_val_loan_income_ratio, file = paste(model_path, "/cross_val_model.rdata", sep=""))
save(merged_2014_model, file = paste(model_path, "/merged_model.rdata", sep=""))

save(model_2014_with_loan_to_value_ratio, file = paste(model_path, "/merged_model_with_loan_to_value.rdata", sep=""))
```

# Save the train and test data for future reference.

```{r}

dir.create(model_data_path, showWarnings = FALSE)
save(train, file = paste(model_data_path, "/train.rdata", sep = ""))
save(test, file = paste(model_data_path, "/test.rdata", sep = ""))

dir.create(model_data_path, showWarnings = FALSE)
save(merged_train, file = paste(model_data_path, "/merged_train.rdata", sep = ""))
save(merged_test, file = paste(model_data_path, "/merged_test.rdata", sep = ""))

dir.create(model_data_path, showWarnings = FALSE)
save(new_train, file = paste(model_data_path, "/train_with_loan_to_value.rdata", sep = ""))
save(new_test, file = paste(model_data_path, "/test_with_loan_to_value.rdata", sep = ""))


```
