---
title: "Naive Bayes for Loansharks (2014-2015 Pennsylvania)"
author: "Omkar Pawar [Please add your names here]" 
date: "4/8/2020"
output: html_notebook
---

## Global setup like working directory, data directory etc should happen here.

```{r}
library(sys)

working_directory <- getwd()

setwd(dirname(dirname(working_directory)))

writeLines("")
getwd()
# The HMDA dataset is loaded from here.
data_dir <- "D:/data"
census_data_dir = "d:/data/census"

# The models are saved in this directory.
model_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "models", sep = "/")
print(model_path)

# The train and test data are saved in this directory
model_data_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "data", sep = "/")
print(model_data_path)

```

## Install required packages.
```{r}
library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)
library(caret)
library(pROC)
library(ROCR)

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/utils.r", sep="/"))

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/model_utils.r", sep="/"))

```

# Load the imputed hmda csv file.
```{r}
#hmda_data_df_ <- fread("/Users/omkarpawar/Desktop/Data/PA/hmda_2014_pa_imputed.csv")

hmda_data_df_ <- fread("/Users/omkarpawar/Desktop/Data/PA/hmda_2014_pa_imputed.csv")
hmda_data_df_ <- as.data.frame(hmda_data_df_)


#hmda_2015 <- fread("/Users/omkarpawar/Desktop/Data/PA/hmda_2015_pa_imputed.csv")
hmda_2015 <- fread("/Users/omkarpawar/Desktop/Data/PA/hmda_2014_pa_imputed.csv")

hmda_2014 <- as.data.frame(hmda_data_df_)
hmda_2014 <- hmda_2014[, !(colnames(hmda_2014) %in% c("Decision"))]
df2015 <- as.data.frame(hmda_2015)

hmda_data_df <- rbind(hmda_2014, df2015)

```


# Explore the dataframe obtained above.
```{r}
head(hmda_data_df, 20)
# Ignore action values of 4 : Application withdrawn by applicant
# 6: Loan purchased by the institution and 2: Application approved but
# not accepted. 
# The remaining action types are 1: Loan originated, 3. Application denied
# by financial institution, 5. File closed for incompleteness and 7. Preapproval
# request denied by financial institution.
hmda_data_df <- hmda_data_df[hmda_data_df$action_taken != 2 & hmda_data_df$action_taken != 4 & hmda_data_df$action_taken != 6, ]

```

# Create dataframe based on selected column list from the source dataframe.
```{r}
hmda_data_df_for_model <- hmda_data_frame_for_model(hmda_data_df)
hmda_data_df_for_model <- process_model_df_columns(hmda_data_df_for_model)
```

# Explore the dataframe with selected columns obtained above.

```{r}
head(hmda_data_df_for_model, 20)

nrow(hmda_data_df_for_model)

length(levels(hmda_data_df_for_model$respondent_id))
length(levels(hmda_data_df_for_model$county_code))

dim(hmda_data_df_for_model)
```

# Train/test split.
```{r}
set.seed(400)

split_dfs <- train_test_split(hmda_data_df_for_model$loan_granted, hmda_data_df_for_model)

train <- split_dfs[[1]]
test <- split_dfs[[2]]
dim(train)
```

# Naive Bayes Model with all Predictors
```{r}
library(e1071)

all_predictors_model <- naiveBayes(loan_granted~., train)
preds <- predict(all_predictors_model,test)
confusionMatrix(preds,test$loan_granted, positive = "yes")
```

```{r}
cm <- confusionMatrix(preds,test$loan_granted, positive = "yes")
cm
draw_confusion_matrix(cm, "Loan Granted", "Loan Denied")
```

# Function to build a Naive Bayes Model.
```{r}
build_nb <- function(formula, df, test){
  model <- naiveBayes(formula, df)
  preds <- predict(model,test)
  cm<-confusionMatrix(preds,test$loan_granted,positive = "yes")
  draw_confusion_matrix(cm,"Loan Granted","Loan Denied")
  precision_recall(cm$table)
  return(model)
}
```

# Model with applicant income and loan amount.
```{r}
nb_2015_income_loan_amount <-
build_nb(as.formula("loan_granted~applicant_race_and_ethnicity+applicant_income_000s+loan_amount_000s+preapproval+co_applicant_present+county_code+minority_population+loan_purpose+tract_to_msamd_income+applicant_sex+property_type+owner_occupancy+agency_code+rate_spread+respondent_id"), train , test)


```


# Model with logs of applicant income and loan amount.
```{r}
nb_2015_log_income_loan_amount <-
build_nb(as.formula("loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+log_loan_amount_000s+preapproval+co_applicant_present+county_code+tract_to_msamd_income+loan_purpose+owner_occupancy+respondent_id+minority_population+rate_spread+agency_code+applicant_sex+property_type"), train , test)
```

# Model with logs of applicant income and loan amount and loan to income ratio

```{r}
nb_2015_ratio_of_loan_to_income_with_log_of_income_and_amount <-
build_nb(as.formula("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+log_loan_amount_000s+applicant_sex+county_code+tract_to_msamd_income+co_applicant_present+agency_code+minority_population+loan_purpose+rate_spread+property_type+loan_to_income_ratio+respondent_id"), train , test)
```


# Model with census data integrated.
```{r}
# https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/counties/asrh/
census_pa_file_path <- "/Users/omkarpawar/Desktop/Data/Census/cc-est2018-alldata-42.csv"

merged_hmda_data_df_for_model <- merge_hmda_df_with_census(census_pa_file_path, hmda_data_df_for_model)

split_dfs <- train_test_split(merged_hmda_data_df_for_model$loan_granted, merged_hmda_data_df_for_model)

merged_train <- split_dfs[[1]]
merged_test <- split_dfs[[2]]

merged_2015_model <-
build_nb(as.formula("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+property_type+log_loan_amount_000s+tract_to_msamd_income+co_applicant_present+agency_code+BLACK_PERCENT+ASIAN_PERCENT+HISPANIC_PERCENT+county_code+respondent_id+applicant_sex+loan_purpose+loan_to_income_ratio"), merged_train , merged_test)

model5_fitted.probs <- predict(merged_2015_model, merged_test, "raw")
```

# Loan to value ratio is an important metric which banks use to decide
# whether a loan should be granted or not. For e.g. the higher the loan
# to value ratio the riskier it is for the bank.
# https://en.wikipedia.org/wiki/Loan-to-value_ratio
# Banks are not required  to report this metric in the HMDA reporting
# law. We use the median of the home values in that county based on
# data available from zillow https://www.zillow.com/research/data/
```{r}
period_list <- c("2015-01", "2015-02", "2015-03", "2015-04",
                 "2015-05", "2015-06", "2015-07", "2015-08",
                 "2015-09", "2015-10", "2015-11", "2015-12")

merged_hmda_with_loan_to_value_ratio <- augment_hmda_data_with_estimated_loan_to_value_ratio_from_zillow("/Users/omkarpawar/Desktop/Data/Secondary/County_Zhvi_AllHomes.csv", hmda_data_df_for_model, "PA", period_list)


head(merged_hmda_with_loan_to_value_ratio)

```

# Train test split on the hmda dataset merged with estimated loan to value
# ratio
```{r}
split_dfs <- train_test_split(merged_hmda_with_loan_to_value_ratio$loan_granted, merged_hmda_with_loan_to_value_ratio)
new_train <- split_dfs[[1]]
new_test <- split_dfs[[2]]
```

## Model with estimated loan to value ratio.
```{r}
nb_2015_with_loan_to_value_ratio <-
build_nb(as.formula("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+log_loan_amount_000s+applicant_sex+county_code+tract_to_msamd_income+co_applicant_present+agency_code+minority_population+loan_to_income_ratio+loan_purpose+rate_spread+property_type+loan_to_value_ratio+respondent_id"), new_train,new_test)

model6_fitted.probs <- predict(nb_2015_with_loan_to_value_ratio, new_test, "raw")
#print_model_performance(model6_fitted.probs, new_test, positive = "yes", negative = "no", threshold = 0.6)

```

# Save the models
```{r}
dir.create(model_path, showWarnings = FALSE)
save(nb_2015_income_loan_amount, file = paste(model_path, "/nb_income_loan_amount.rdata", sep=""))
save(nb_2015_log_income_loan_amount, file = paste(model_path, "/nb_log_income_loan_amount.rdata", sep=""))
save(nb_2015_ratio_of_loan_to_income_with_log_of_income_and_amount, file = paste(model_path, "/nb_loan_income_ratio_with_log_of_income_and_amount.rdata", sep=""))
save(merged_hmda_with_loan_to_value_ratio, file = paste(model_path, "/merged_model.rdata", sep=""))

save(nb_2015_with_loan_to_value_ratio, file = paste(model_path, "/nb_merged_model_with_loan_to_value.rdata", sep=""))
```

# Save the train and test data for future reference.

```{r}
dir.create(model_data_path, showWarnings = FALSE)
save(train, file = paste(model_data_path, "/train.rdata", sep = ""))
save(test, file = paste(model_data_path, "/test.rdata", sep = ""))

dir.create(model_data_path, showWarnings = FALSE)
save(merged_train, file = paste(model_data_path, "/merged_train.rdata", sep = ""))
save(merged_test, file = paste(model_data_path, "/merged_test.rdata", sep = ""))

dir.create(model_data_path, showWarnings = FALSE)
save(new_train, file = paste(model_data_path, "/train_with_loan_to_value.rdata", sep = ""))
save(new_test, file = paste(model_data_path, "/test_with_loan_to_value.rdata", sep = ""))


```
