---
title: "Logistic Regression for Loan Sharks (PA 2007 and 2008)"
author: "Anantanarayanan G Iyengar"
date: "4/25/2020"
output: html_notebook
---

## We tried the following glm models. The predictors chosen were picked
## from our EDA analysis and feature engineering work and some background
## knowledge.
# Model 1. Log of loan and applicant income.
# Model 2. Log of loan amount and loan to income ratio.

## Results.
# Model 1, Model 2 performed similarly. The criterion was AUC 
# curve, Mcfadden R square value and confusion matrix results on test data.

# Model 2 was chosen as banks do use the loan to income ratio as one of their
# criterion. 

## Key observation
# The AUC curve for our models ranges between 83 to 84%. The AUC curve can
# be thought of as a classification-threshold-invariant. It measures the #
# quality of the model's predictions irrespective of the threshold value
# https://developers.google.com/machine-learning/crash-course/classification/roc-and-auc.
# It is surprising that the models for 2007 and 2008 have higher AUC curves
# when compared with those for 2014-2017. We feel that this is likely due to
# the lower lending standards which were rampant during that period.


## Global setup like working directory, data directory etc should happen here.

```{r}
library(sys)

working_directory <- getwd()

setwd(dirname(dirname(working_directory)))

writeLines("")
getwd()
# The HMDA dataset is loaded from here.
data_dir <- "D:/data"
census_data_dir = "d:/data/census"

# The models are saved in this directory.
model_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "models", sep = "/")
print(model_path)

# The train and test data are saved in this directory
model_data_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "data", sep = "/")
print(model_data_path)

```

## Install required packages.
```{r}
# https://stackoverflow.com/questions/4090169/elegant-way-to-check-for-missing-packages-and-install-them
list_of_packages <- c("mlbench", "corrplot", "rvest", "tidyr", "stringr", "dplyr", "lubridate", "data.table", "mice", "scales", "naniar", "rpart", "rpart.plot", "caret", "pROC", "ROCR")
new.packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]

if (length(new.packages)) {
  print("Installing packages\n")
  install.packages(new.packages())
}

library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)
library(caret)
library(pROC)
library(ROCR)

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/utils.r", sep="/"))

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/model_utils.r", sep="/"))

options(scipen = 999)

```

# Load the imputed hmda csv file.
```{r}
hmda_data_df_2007 <- fread(paste(data_dir, "/HMDA/2007/hmda_2007_pa_imputed.csv", sep=""))
hmda_data_df_2007 <- as.data.frame(hmda_data_df_2007)

hmda_data_df_2008 <- fread(paste(data_dir, "/HMDA/2008/hmda_2008_pa_imputed.csv", sep=""))
hmda_data_df_2008 <- as.data.frame(hmda_data_df_2008)
```

# Visualize the loaded data frames.
```{r}

head(hmda_data_df_2007)

head(hmda_data_df_2008)


```


# Create dataframe based on selected column list from the source dataframe.
```{r}
hmda_data_df_for_model_2007 <- hmda_data_frame_for_model(hmda_data_df_2007)
hmda_data_df_for_model_2007 <- process_model_df_columns(hmda_data_df_for_model_2007)

hmda_data_df_for_model_2008 <- hmda_data_frame_for_model(hmda_data_df_2008)
hmda_data_df_for_model_2008 <- process_model_df_columns(hmda_data_df_for_model_2008)
```

# Train/test split.
```{r}
set.seed(400)

split_dfs <- train_test_split(hmda_data_df_for_model_2007$loan_granted, hmda_data_df_for_model_2007)

train_2007 <- split_dfs[[1]]
test_2007 <- split_dfs[[2]]

split_dfs <- train_test_split(hmda_data_df_for_model_2008$loan_granted, hmda_data_df_for_model_2008)

train_2008 <- split_dfs[[1]]
test_2008 <- split_dfs[[2]]

```

# Null model
```{r}
null_model_2007 <- build_and_evaluate_hmda_null_model("loan_granted~1", train_2007, test_2007)

null_model_2008 <- build_and_evaluate_hmda_null_model("loan_granted~1", train_2008, test_2008)

```

# Model with logs of applicant income and loan amount for 2007
```{r}
model_2007_log_income_loan_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+log_loan_amount_000s+preapproval+co_applicant_present+county_code+tract_to_msamd_income+loan_purpose+owner_occupancy+respondent_id+minority_population+rate_spread+agency_code+applicant_sex+property_type", train_2007)

model_summary_and_plot(model_2007_log_income_loan_amount, train_2007)
mcfadden_r_square(model_2007_log_income_loan_amount, null_model_2007)

```

# Model with logs of loan amount and loan to income ratio for 2007
```{r}
model_2007_ratio_of_loan_to_income_with_log_of_income_and_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+log_loan_amount_000s+preapproval+co_applicant_present+county_code+tract_to_msamd_income+loan_purpose+owner_occupancy+respondent_id+minority_population+rate_spread+agency_code+applicant_sex+property_type+loan_to_income_ratio", train_2007)

model_summary_and_plot(model_2007_ratio_of_loan_to_income_with_log_of_income_and_amount, train_2007)
mcfadden_r_square(model_2007_ratio_of_loan_to_income_with_log_of_income_and_amount, null_model_2007)

```
# Model with logs of applicant income and loan amount for 2008
```{r}
model_2008_log_income_loan_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+log_loan_amount_000s+preapproval+co_applicant_present+county_code+tract_to_msamd_income+loan_purpose+owner_occupancy+respondent_id+minority_population+rate_spread+agency_code+applicant_sex+property_type", train_2008)

model_summary_and_plot(model_2008_log_income_loan_amount, train_2008)
mcfadden_r_square(model_2008_log_income_loan_amount, null_model_2008)

```


# Model with logs of loan amount and loan to income ratio for 2008
```{r}
model_2008_ratio_of_loan_to_income_with_log_of_income_and_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+log_loan_amount_000s+preapproval+co_applicant_present+county_code+tract_to_msamd_income+loan_purpose+owner_occupancy+respondent_id+minority_population+rate_spread+agency_code+applicant_sex+property_type+loan_to_income_ratio", train_2008)

model_summary_and_plot(model_2008_ratio_of_loan_to_income_with_log_of_income_and_amount, train_2008)
mcfadden_r_square(model_2008_ratio_of_loan_to_income_with_log_of_income_and_amount, null_model_2008)

```

# Model performance for 2007

```{r}
library(ROCR)
library(pROC)
model1_fitted.probs <- predict(model_2007_log_income_loan_amount, test_2007, type = "response")

model2_fitted.probs <- predict(model_2007_ratio_of_loan_to_income_with_log_of_income_and_amount, test_2007, type = "response")

writeLines("")
writeLines("Model 1 2007 performance")
print_model_performance(model1_fitted.probs, test_2007, positive = "yes", negative = "no", threshold = 0.3)

writeLines("")
writeLines("Model 2 2007 performance")
print_model_performance(model2_fitted.probs, test_2007, positive = "yes", negative = "no", threshold = 0.3)

```

# Model performance for 2008

```{r}
library(ROCR)
library(pROC)
model3_fitted.probs <- predict(model_2008_log_income_loan_amount, test_2008, type = "response")

model4_fitted.probs <- predict(model_2008_ratio_of_loan_to_income_with_log_of_income_and_amount, test_2008, type = "response")

writeLines("")
writeLines("Model 1 2008 performance")
print_model_performance(model3_fitted.probs, test_2008, positive = "yes", negative = "no", threshold = 0.3)

writeLines("")
writeLines("Model 2 2008 performance")
print_model_performance(model4_fitted.probs, test_2008, positive = "yes", negative = "no", threshold = 0.3)

```


# Save the models
```{r}
dir.create(model_path, showWarnings = FALSE)
save(model_2007_log_income_loan_amount, file = paste(model_path, "/model_2007_log_income_loan_amount.rdata", sep=""))
save(model_2007_ratio_of_loan_to_income_with_log_of_income_and_amount, file = paste(model_path, "/model_2007_loan_income_ratio_with_log_of_income_and_amount.rdata", sep=""))

save(model_2008_log_income_loan_amount, file = paste(model_path, "/model_2008_log_income_loan_amount.rdata", sep=""))
save(model_2008_ratio_of_loan_to_income_with_log_of_income_and_amount, file = paste(model_path, "/model_2008_loan_income_ratio_with_log_of_income_and_amount.rdata", sep=""))

```

# Save the train and test data for future reference.

```{r}

dir.create(model_data_path, showWarnings = FALSE)
save(train_2007, file = paste(model_data_path, "/train_2007.rdata", sep = ""))
save(test_2007, file = paste(model_data_path, "/test_2007.rdata", sep = ""))

dir.create(model_data_path, showWarnings = FALSE)
save(train_2008, file = paste(model_data_path, "/train_2008.rdata", sep = ""))
save(test_2008, file = paste(model_data_path, "/test_2008.rdata", sep = ""))

```