---
title: "Project Experiment for Illinois Home Loans for 2017"
output: html_notebook
---

## Global setup like working directory, data directory etc should happen here.
```{r}
# Please change the working directory as needed.
working_dir <- "/Users/omkarpawar/Downloads"
setwd(working_dir)

data_dir <- "D:/data/HMDA"
```

## Load essential Libraries
```{r}
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(corrplot)
library(data.table)
library(mice)
library(rstudioapi)    
#library(naniar)

```

## Load Data
```{r}
hmda_data_il <- fread(paste(working_dir, "/hmda_2017_il_all-records_labels.csv", sep = ""))
```
## Data Exploration

Print head rows of the dataframe and the column names
```{r}
hmda_il_df <- as.data.frame(hmda_data_il)
head(hmda_il_df)
```


```{r}
colnames(hmda_il_df)
```


```{r}
summary(hmda_il_df)
df_copy<-hmda_il_df
```


```{r}
dim(hmda_il_df)
```


```{r}
glimpse(hmda_il_df)
```

## Find and Handle missing values
Look for NA
```{r}
sapply(hmda_il_df, function(x) sum(is.na(x)))
```
Check for any empty strings

```{r}
sapply(hmda_il_df, function(x) sum(x == ""))
```


```{r}
sapply(hmda_il_df, function(x) sum(x == NULL))
```

## Race and Ethnicity columns and insights
Race
```{r}
unique(hmda_il_df$applicant_race_name_1)
```
```{r}
unique(hmda_il_df$applicant_ethnicity_name)
```


```{r}
table(hmda_il_df$applicant_race_name_1)
```

```{r}
hmda_il_df$applicant_race_and_ethnicity <- NA
hmda_il_df$co_applicant_race_and_ethnicity <- NA

hmda_il_df$applicant_race_and_ethnicity <- ifelse(hmda_il_df$applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_il_df$applicant_race_name_1)

hmda_il_df$co_applicant_race_and_ethnicity <- ifelse(hmda_il_df$co_applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_il_df$co_applicant_race_name_1)


head(hmda_il_df)
```


```{r}
count_racename <- as.data.frame(table(hmda_il_df$applicant_race_and_ethnicity))
colnames(count_racename)<-c("Race_and_Ethnicity","Count")

p<-ggplot(data=count_racename, aes(x=reorder(Race_and_Ethnicity,-Count), y=Count)) + geom_bar(stat="identity",fill="steelblue")+geom_text(aes(label=Count), size=2.5)
p + coord_flip() +labs(title = "Distribution of applicants by Race and Ethnicity", y="Number of Applicants",x="Race and Ethnicity")

#text(x = p, y = count_racename$Count, label = count_racename$Count, pos = 3, cex = 0.8, col = "red")
```


```{r}
count_racename[order(count_racename$Count),]
```


```{r}
library(plotrix)
#pie3D(count_racename$Count,labels = count_racename$Race_and_Ethnicity,explode=0.1)
```


```{r}
hmda_origination_status_df <- hmda_il_df[hmda_il_df$action_taken == "1", ]

head(hmda_origination_status_df)

ggplot(hmda_origination_status_df, aes(x=applicant_income_000s)) + geom_histogram(binwidth = 10)

max(hmda_origination_status_df$applicant_income_000s)-min(hmda_origination_status_df$applicant_income_000s)

hist(hmda_origination_status_df$applicant_income_000s,xlim = c(0,100))

class(hmda_origination_status_df)

sum(is.na(hmda_origination_status_df$applicant_income_000s))/length(hmda_origination_status_df$applicant_income_000s)*100

hmda_origination_status_df$applicant_income_000s[is.na(hmda_origination_status_df$applicant_income_000s)]<-median(hmda_origination_status_df$applicant_income_000s,na.rm = TRUE)

length(hmda_origination_status_df$applicant_income_000s[hmda_origination_status_df$applicant_income_000s>1000])
```
## Income of applicants
```{r}
summary(hmda_il_df$applicant_income_000s)

```
Here it can be observed that the income of applicants is spreadout a lot. We have applicants with income ranging from 1 to 147759 in thousands. Also, missing values sum upto 65583.

Replace missing values in applicants income with median.
```{r}
hmda_il_df$applicant_income_000s[is.na(hmda_il_df$applicant_income_000s)]<-median(hmda_il_df$applicant_income_000s,na.rm = TRUE)
```
Find the income distribution based on race and ethnicity
```{r}
p <- ggplot(group_by(hmda_il_df,applicant_race_and_ethnicity), aes(x=applicant_race_and_ethnicity, y=applicant_income_000s)) + 
  geom_boxplot()
p +coord_flip()
```
In the boxplot, the category where people didn't provide the race and ethnicity data has most extreme values. 

### These are probably outliers and we need to deal with them
```{r}
p <- ggplot(group_by(hmda_origination_status_df,applicant_race_and_ethnicity), aes(x=applicant_race_and_ethnicity, y=applicant_income_000s)) + 
  geom_boxplot()
p +coord_flip()
```

```{r}
unique(hmda_il_df$action_taken_name)
```
Here we are trying to find the distribution of loans that are denied over race and ethnicity.
## Create a new column where we classify the action taken as loan is denied or not.
```{r}
hmda_il_df$denied
hmda_il_df$denied<-NA

hmda_il_df$denied[hmda_il_df$action_taken_name == "Application denied by financial institution" | hmda_il_df$action_taken_name == "Preapproval request denied by financial institution"]<- "Denied"

hmda_il_df$denied[is.na(hmda_il_df$denied)]<-"Approved"
unique(hmda_il_df$denied)

application_denied<-sum(hmda_il_df$denied == "Denied")
application_originated<-sum(hmda_il_df$denied == "Approved")
application_denied/length(hmda_il_df$action_taken)
group_by(hmda_il_df$action_taken)

ggplot(hmda_il_df,aes(hmda_il_df$applicant_race_and_ethnicity, ..count..)) +geom_bar(aes(fill= hmda_il_df$denied))
library(ggplot2)
class(hmda_il_df$denied)
hmda_il_df$denied - as.character(hmda_il_df$denied)
p<-ggplot(hmda_il_df , aes (x = applicant_race_name_1 , fill = denied,)) + geom_bar(position = "dodge")
p + coord_flip()
hmda_il_df$denied

?summarise_at(group_by(hmda_il_df,applicant_race_name_1),vars(denied),func())
```
