---
title: "New_Analysis_2016_IL"
output: html_document
---
```{r}
#install.packages('summarytools')
#install.packages("caroline")
library(summarytools)
library(tidyverse)
library(plyr)
library(rpart)
library(rpart.plot)
library(caTools)
library(caret)
library(DT)
library(dplyr)
library(caroline)
library(Hmisc)
rm(list=ls())

homeMortgage = read_csv("C:/Users/bhuvn/Downloads/HDMA_2016_IL.csv")
dim(homeMortgage)

d<-describe(homeMortgage)
d
r<-dfSummary(homeMortgage)
r
```


Variables with all observations missing:
[1] applicant_race_name_3    applicant_race_name_4    applicant_race_name_5    applicant_race_5        
[5] co_applicant_race_name_2 co_applicant_race_name_3 co_applicant_race_name_4 co_applicant_race_4     
[9] co_applicant_race_name_5 co_applicant_race_5 
```{r}
drops <- c("applicant_race_name_3","applicant_race_name_4","applicant_race_name_5","applicant_race_5",
           "co_applicant_race_name_2","co_applicant_race_name_3","co_applicant_race_name_4","co_applicant_race_4",
           "co_applicant_race_name_5","co_applicant_race_5")
homeMortgage <- homeMortgage[ , !(names(homeMortgage) %in% drops)]
dim(homeMortgage)
```

Variables with above 98% observations missing:
```{r}
drops <- c("rate_spread","denial_reason_3","denial_reason_name_3","denial_reason_2","denial_reason_name_2",
           "co_applicant_race_3","co_applicant_race_2","applicant_race_4","applicant_race_3",
           "applicant_race_2","applicant_race_name_2")
homeMortgage <- homeMortgage[ , !(names(homeMortgage) %in% drops)]
dim(homeMortgage)
```

Variables with No unique values (Not important factors)
```{r}
drops <- c("as_of_year","state_name","state_abbr","state_code")
homeMortgage <- homeMortgage[ , !(names(homeMortgage) %in% drops)]
dim(homeMortgage)
```

Removing the column name containing the codes for the columns with string values.
Not required as they are a repetion of the corresponding column
```{r}
drops <- c("agency_code","loan_type","property_type","loan_purpose","owner_occupancy","preapproval","action_taken",
           "msamd","county_code","applicant_ethnicity","co_applicant_ethnicity","applicant_race_1","co_applicant_race_1",
           "applicant_sex","co_applicant_sex","purchaser_type","denial_reason_1","hoepa_status","lien_status","edit_status")
homeMortgage <- homeMortgage[ , !(names(homeMortgage) %in% drops)]
dim(homeMortgage)
```

Respondent Information features like Respondant ID, Agency, and Edit Status are not relevant to our business problem.

```{r}
drops <- c("respondent_id","agency_name","edit_status_name")
homeMortgage <- homeMortgage[ , !(names(homeMortgage) %in% drops)]
dim(homeMortgage)
```


print(describe(homeMortgage)) #Out of 30 predictors, we are geting missing values in 11 predictors
Counting Number of NA in each row

```{r}
homeMortgage$na_count <- apply(homeMortgage, 1, function(x) sum(is.na(x)))
print(count(homeMortgage, "na_count"))
```

Removing the rows with NA counts per row more than 9


```{r}
homeMortgage<-homeMortgage[!(homeMortgage$na_count==11 | homeMortgage$na_count==10 | homeMortgage$na_count==9),]
dim(homeMortgage)
```

Denial reason name is having too many missing value and it is not relevant
```{r}
drops <- c("denial_reason_name_1")
homeMortgage <- homeMortgage[ , !(names(homeMortgage) %in% drops)]
dim(homeMortgage)
```

Appling NA count method again
```{r}
homeMortgage$na_count <- apply(homeMortgage, 1, function(x) sum(is.na(x)))
print(count(homeMortgage, "na_count"))
```

```{r}
lookup_df <- as.data.frame(summarise_at(group_by(homeMortgage,county_name),
                                     vars(hud_median_family_income),
                                     funs(mean(.,na.rm=TRUE))))
lookup_df
dim(lookup_df)
dim(homeMortgage)
```

```{r}
homeMortgage$applicant_income_000s[homeMortgage$applicant_income_000s==""] <- "NA"
dim(homeMortgage)
a<-count(is.na(homeMortgage$applicant_income_000s))
a
```

Replacing the missing Income values with Median family income in county
```{r}
homeMortgage$applicant_income_000s = ifelse(is.na(homeMortgage$applicant_income_000s),
                                            paste((homeMortgage$hud_median_family_income/1000)),
                                            paste(homeMortgage$applicant_income_000s));homeMortgage

count(is.na(homeMortgage$applicant_income_000s))
```

Appling NA count method again
```{r}
homeMortgage$na_count <- apply(homeMortgage, 1, function(x) sum(is.na(x)))
print(count(homeMortgage, "na_count"))

dfSummary(homeMortgage)
```


Removing rows with empty census tract number.
```{r}
new_DF <- homeMortgage[is.na(homeMortgage$census_tract_number),]
table(new_DF$applicant_race_name_1)
dim(homeMortgage)
homeMortgage<-homeMortgage[!(is.na(homeMortgage$census_tract_number)),]
dim(homeMortgage)
```

Appling NA count method again
```{r}
homeMortgage$na_count <- apply(homeMortgage, 1, function(x) sum(is.na(x)))
print(count(homeMortgage, "na_count"))
dfSummary(homeMortgage)
```

```{r}
new1_DF <- homeMortgage[is.na(homeMortgage$tract_to_msamd_income),]
new2_DF <- homeMortgage[is.na(homeMortgage$population),]
new1_DF %>% select(tract_to_msamd_income, hud_median_family_income,county_name,applicant_income_000s)
new2_DF %>% select(county_name,population,minority_population,census_tract_number,msamd_name)

homeMortgage$tract_to_msamd_income = ifelse(is.na(homeMortgage$tract_to_msamd_income),
                                            paste(((as.numeric(homeMortgage$applicant_income_000s)*1000)/homeMortgage$hud_median_family_income)*100),
                                            paste(homeMortgage$tract_to_msamd_income));homeMortgage

count(is.na(homeMortgage$tract_to_msamd_income))

dfSummary(homeMortgage)

new3_DF <- homeMortgage[is.na(homeMortgage$number_of_owner_occupied_units),]
new3_DF %>% select(number_of_owner_occupied_units, census_tract_number,msamd_name)
```





