---
title: "CSP571_Project_Experimentation for Pennsylvania conventional home loans for 2014"
authors: "Anantanarayanan G Iyengar, [Please add your names here]"
date: "2/17/2020"
output: html_notebook
---

## Global setup like working directory, data directory etc should happen here.
```{r}
# Please change the working directory as needed.
working_dir <- "D:/illinois/csp571/csp-571-02-final-project"
setwd(working_dir)

data_dir <- "D:/data/HMDA"
```

## Install required packages.
```{r}
# https://stackoverflow.com/questions/4090169/elegant-way-to-check-for-missing-packages-and-install-them
list_of_packages <- c("mlbench", "corrplot", "rvest", "tidyr", "stringr", "dplyr", "lubridate", "data.table", "mice", "scales", "naniar", "rpart", "rpart.plot", "caret")
new.packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]

if (length(new.packages)) {
  print("Installing packages\n")
  install.packages(new.packages())
}

library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)

source(paste(dirname(dirname(rstudioapi::getActiveDocumentContext()$path)), "utils/utils.r", sep="/"))

```

## Load data file.
```{r}
hmda_data_pa <- fread("hmda_2014_pa_all-records_labels.csv")
```

## Data analysis section 1. Print column names and first 10 rows of dataset.
```{r}
hmda_data_pa_df <- as.data.frame(hmda_data_pa)

# Filter to include conventional loans only.
hmda_data_pa_df <- hmda_data_pa_df[hmda_data_pa_df$loan_type == "1", ]

colnames(hmda_data_pa_df)

writeLines("")

summary(hmda_data_pa_df)

head(hmda_data_pa_df, 10)
```

## Data analysis section 2. Print glimpse of dataset i.e a vertical preview of the dataset.
```{r}
dim(hmda_data_pa_df)
writeLines("Glimpse of hmda dataset for PA")
glimpse(hmda_data_pa_df)

```

## Data analysis section 2. Print structure of dataset.

```{r}
str(hmda_data_pa_df)

```

## Data Analysis Section 3. Check for missing values in the dataset
```{r}
writeLines("Checking for missing values with NA")
sapply(hmda_data_pa_df, function(x) sum(is.na(x)))

writeLines("Checking for missing values with empty strings")
sapply(hmda_data_pa_df, function(x) sum(x == ""))

writeLines("Checking for missing values with null")
sapply(hmda_data_pa_df, function(x) sum(x == NULL))

```

## Look at values in the ethnicity and race columns

```{r}
library(janitor)

writeLines("Application ethnicity values")
unique(hmda_data_pa_df$applicant_ethnicity_name)

writeLines("Application race name 1 values")
unique(hmda_data_pa_df$applicant_race_1)
unique(hmda_data_pa_df$applicant_race_name_1)

hmda_data_pa_df %>% filter(applicant_ethnicity_name == "Hispanic or Latino") %>%
  group_by(applicant_race_name_1) %>% 
           count() %>%
           ungroup() %>%
           replace(is.na(.), 0) %>% 
           adorn_totals(c("col")) %>% 
           arrange(-Total)
```
## Add a new column applicant_race_and_ethnicity and group all applicants
## with ethnicity as Hispanic or Latino as Hispanic or Latino in this column.
## For everyone else, this column gets values from the applicant_race_name_1 
## column

```{r}
hmda_data_pa_df$applicant_race_and_ethnicity <- NA
hmda_data_pa_df$co_applicant_race_and_ethnicity <- NA

hmda_data_pa_df$applicant_race_and_ethnicity <- ifelse(hmda_data_pa_df$applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_data_pa_df$applicant_race_name_1)

hmda_data_pa_df$co_applicant_race_and_ethnicity <- ifelse(hmda_data_pa_df$co_applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_data_pa_df$co_applicant_race_name_1)

#hmda_data_pa_df$applicant_race_and_ethnicity <- hmda_data_pa_df$applicant_race_and_ethnicity

#hmda_data_pa_df$co_applicant_race_and_ethnicity <- hmda_data_pa_df$co_applicant_race_and_ethnicity

head(hmda_data_pa_df)

unique(hmda_data_pa_df$applicant_race_and_ethnicity)
```

## Graph mortgage distribution by applicant race and ethinicity.
```{r}
mortgage_by_race_and_ethnicity = hmda_data_pa_df %>% group_by(applicant_race_and_ethnicity) %>%
  summarise(EthnicityCount = n()) %>%
  arrange(desc(EthnicityCount))

graph_by_enthicity(mortgage_by_race_and_ethnicity)
```

# Graph which applicant races and ethnicities have the largest proportion of loans
# in various stages. These include origination status, denied status, etc.

```{r}
mortgage_status_by_race_and_ethnicity <- hmda_data_pa_df %>% group_by(action_taken_name, applicant_race_and_ethnicity) %>%
  summarise(ActionCount = n()) %>%
  arrange(desc(ActionCount))

mortgage_status_aggregated_by_race_and_ethnicity  = inner_join(mortgage_status_by_race_and_ethnicity, mortgage_by_race_and_ethnicity) %>% mutate(percentage = (ActionCount / EthnicityCount) * 100)

graph_application_race_proportion_of_loans(mortgage_status_aggregated_by_race_and_ethnicity)

```

# Applicant income histograms.

```{r}
hmda_origination_status_df <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1", ]
graph_applicant_income_histogram(hmda_origination_status_df, "Applicant income distribution for originated loans")
```

# Graph median income for originated loans.
```{r}
hmda_origination_status_df <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1", ]

head(hmda_origination_status_df)

hmda_origination_status_df %>% ggplot(aes(as.numeric(hud_median_family_income))) +
geom_histogram(binwidth = 1000,, fill=c("blue")) + labs(x = "Median Income", y = "Applicant Count", title = "Median Income Distribution for Area for Originated Loans") + theme_bw()
```

## Graph loan distribution by county.

```{r}
mortgage_distribution_by_counties <- hmda_data_pa_df %>%
  filter(!is.na(county_name)) %>%
  group_by(county_name) %>%
  summarise(CountLoans = n() ) %>%
  mutate(percentage = ( CountLoans / sum(CountLoans) ) * 100 ) %>%
  mutate(county_name = reorder(county_name, percentage)) %>%
  arrange(desc(percentage)) %>%
  head(20)

graph_distribution_by_county(mortgage_distribution_by_counties)

originated_mortgage_distribution_by_counties <- hmda_origination_status_df %>%
  filter(!is.na(county_name)) %>%
  group_by(county_name) %>%
  summarise(CountLoans = n() ) %>%
  mutate(percentage = ( CountLoans / sum(CountLoans) ) *100 ) %>%
  mutate(county_name = reorder(county_name, percentage)) %>%
  arrange(desc(percentage)) %>%
  head(20)

graph_distribution_by_county(originated_mortgage_distribution_by_counties)

```

## Graph home loan application distribution for the top 4 counties in the above
## chart by applicant_race_1
```{r}

county_names <- c("Allegheny County", "Philadelphia County", "Montgomery County", "Bucks County")

for (county_name in county_names) {
  
  hmda_data_county_df <- hmda_data_pa_df[hmda_data_pa_df$county_name == county_name, ]
  
  mortgage_by_race_county <- hmda_data_county_df %>% group_by(applicant_race_name_1) %>%
    summarise(RaceCount = n()) %>% arrange(desc(RaceCount))
  
  print(graph_mortgage_distribution_by_race1(mortgage_by_race_county))
}

```


## Graph income distribution for Whites and African Americans per county for the
## top 4 counties above.
```{r}

for (county_name in county_names) {
  hmda_origination_status_df_by_county_white <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1" & hmda_data_pa_df$county_name == county_name & hmda_data_pa_df$applicant_race_name_1 == "White", ]
  print(graph_applicant_income_histogram(hmda_origination_status_df_by_county_white, "Income distribution for Whites"))

  hmda_origination_status_df_by_county_african_american <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1" & hmda_data_pa_df$county_name == county_name & hmda_data_pa_df$applicant_race_name_1 == "Black or African American", ]
  print(graph_applicant_income_histogram(hmda_origination_status_df_by_county_african_american, "Income distribution for African Americans"))
}

```


## Graph home loan application distribution for the top 4 counties in the above
## chart by applicant_race_and_ethnicity
```{r}

county_names <- c("Allegheny County", "Philadelphia County", "Montgomery County", "Bucks County")

for (county_name in county_names) {
  
  hmda_data_county_df <- hmda_data_pa_df[hmda_data_pa_df$county_name == county_name, ]
  
  mortgage_by_race_county <- hmda_data_county_df %>% group_by(applicant_race_and_ethnicity) %>%
    summarise(RaceCount = n()) %>% arrange(desc(RaceCount))
  
  print(graph_mortgage_distribution_by_race_and_ethnicity(mortgage_by_race_county))
}

```

# Graph which communities have the largest proportion of loans in various stages.
# for the top 4 counties listed above.
# These include origination status, denied status, etc.

```{r}

for (county_name in county_names) {
  
  hmda_data_county_df <- hmda_data_pa_df[hmda_data_pa_df$county_name == county_name, ]
  
  mortgage_by_race1_county <- hmda_data_county_df %>% group_by(applicant_race_and_ethnicity) %>%
    summarise(RaceCount = n()) %>% arrange(desc(RaceCount))

  mortgage_status_by_race1_by_county <- hmda_data_county_df %>% group_by(action_taken_name, applicant_race_and_ethnicity) %>%
    summarise(ActionCount = n()) %>%
    arrange(desc(ActionCount))
  
  mortgage_status_aggregated_by_race1_by_county  = inner_join(mortgage_status_by_race1_by_county, mortgage_by_race1_county) %>% mutate(percentage = (ActionCount / RaceCount) * 100)
  
  print(graph_application_race_and_ethnicity_proportion_of_loans(mortgage_status_aggregated_by_race1_by_county))
}

```

# Visualize missing variables and impute as needed.
```{r}
library(mice)
library(naniar)

# https://cran.r-project.org/web/packages/naniar/vignettes/naniar-visualisation.html
gg_miss_upset(hmda_data_pa_df)
hmda_data_pa_df$loan_to_income_ratio
# https://www.rdocumentation.org/packages/mice/versions/3.8.0/topics/mice.impute.cart
#Selecting columns which will be used for modelling. 
#Takes time to impute missing values which are not needed.
keep <- c("applicant_race_and_ethnicity", "applicant_income_000s", "co_applicant_race_and_ethnicity", "co_applicant_sex", "county_code", "hoepa_status", "lien_status", "msamd", "owner_occupancy", "preapproval", "property_type", "purchaser_type", "loan_amount_000s", "county_name", "applicant_sex", "hud_median_family_income", "tract_to_msamd_income",  "agency_code", "minority_population")
hmda_data_pa_df_imputed$loan_gr
hmda_data_pa_df_imputed <- mice(hmda_data_pa_df[keep], m=1, maxit=2, meth='cart',seed=500)
hmda_data_pa_df_imputed <- mice::complete(hmda_data_pa_df_imputed)
sapply(hmda_data_pa_df_imputed,function (x) sum(is.na(x)))
```

## More analysis on the imputed dataset.
```{r}

summary(hmda_data_pa_df_imputed)

gg_miss_upset(hmda_data_pa_df_imputed)

```

# Additional analysis on the hmda dataset. correlation matrix, plots, etc.
```{r}
# https://stackoverflow.com/questions/20637360/convert-all-data-frame-character-columns-to-factors
hmda_data_pa_df$loan_to_income_ratio <- hmda_data_pa_df$loan_amount_000s / hmda_data_pa_df$applicant_income_000s

hmda_data_pa_df[sapply(hmda_data_pa_df, is.character)] <- lapply(hmda_data_pa_df[sapply(hmda_data_pa_df, is.character)], 
                                       as.factor)

hmda_data_pa_df_for_correlation <- as.data.frame(lapply(hmda_data_pa_df, as.integer))

#head(hmda_data_pa_df_for_correlation[, c("applicant_income_000s", "loan_amount_000s")])

head(hmda_data_pa_df_for_correlation)

corr_simple(hmda_data_pa_df_for_correlation)

corrplot(cor(hmda_data_pa_df_for_correlation[, c("applicant_income_000s", "loan_amount_000s")], use = "na.or.complete"))

```


# Additional analysis on the hmda imputed dataset. correlation plots, etc.
```{r}
# hmda_data_pa_df_imputed <- hmda_data_pa_df;
# https://stackoverflow.com/questions/20637360/convert-all-data-frame-character-columns-to-factors
hmda_data_pa_df_imputed$loan_to_income_ratio <- hmda_data_pa_df_imputed$loan_amount_000s / hmda_data_pa_df_imputed$applicant_income_000s

hmda_data_pa_df_imputed[sapply(hmda_data_pa_df_imputed, is.character)] <- lapply(hmda_data_pa_df_imputed[sapply(hmda_data_pa_df_imputed, is.character)], 
                                       as.factor)

hmda_data_pa_df_imputed_for_correlation <- as.data.frame(lapply(hmda_data_pa_df_imputed, as.integer))

head(hmda_data_pa_df_imputed_for_correlation[, c("applicant_income_000s", "loan_amount_000s")])

corr_simple(hmda_data_pa_df_imputed_for_correlation)

corrplot(cor(hmda_data_pa_df_imputed_for_correlation[, c("applicant_income_000s", "loan_amount_000s")], use = "na.or.complete"))

```

# Preparation for inital model. Selecting columns. 

```{r}
data_frame_for_model <- function(input_df) {
  # Added based on intuition. Please update as needed.
  # msamd -> Metropolitan Statistical Area Division code.
  # https://cfpb.github.io/api/hmda/fields.html
  initial_columns_for_model <- c("applicant_race_and_ethnicity", "applicant_income_000s", "co_applicant_race_and_ethnicity", "co_applicant_sex", "county_code", "hoepa_status", "lien_status", "msamd", "owner_occupancy", "preapproval", "property_type", "purchaser_type", "loan_amount_000s", "county_name", "applicant_sex", "hud_median_family_income", "tract_to_msamd_income", "loan_to_income_ratio", "agency_code", "minority_population")

  # https://stackoverflow.com/questions/10085806/extracting-specific-columns-from-a-data-frame
  df_for_model <- input_df %>% select(initial_columns_for_model)
  df_for_model$loan_granted <- input_df$action_taken == 1 | input_df$action_taken == 6
  return (df_for_model)
}

```

# Create dataframe based on selected column list from the source dataframe.
```{r}
hmda_data_pa_df_imputed$action_taken <- hmda_data_pa_df$action_taken
hmda_data_pa_df_imputed_for_model <- data_frame_for_model(hmda_data_pa_df_imputed)

```

```{r}

unique(hmda_data_pa_df_imputed$co_applicant_race_and_ethnicity)

hmda_data_pa_df_imputed_for_model$co_applicant_present <- NA

hmda_data_pa_df_imputed_for_model$co_applicant_present <- ifelse( hmda_data_pa_df_imputed_for_model$co_applicant_race_and_ethnicity == "White" | hmda_data_pa_df_imputed_for_model$co_applicant_race_and_ethnicity == "Black or African American" | hmda_data_pa_df_imputed_for_model$co_applicant_race_and_ethnicity == "Hispanic or Latino" | hmda_data_pa_df_imputed_for_model$co_applicant_race_and_ethnicity == "Asian" | hmda_data_pa_df_imputed_for_model$co_applicant_race_and_ethnicity == "Native Hawaiian or Other Pacific Islander" | hmda_data_pa_df_imputed_for_model$co_applicant_race_and_ethnicity == "American Indian or Alaska Native", 1, 0)

#hmda_data_pa_df_imputed_for_model$co_applicant_present <- as.factor(hmda_data_pa_df_imputed_for_model$co_applicant_present)

nrow(hmda_data_pa_df_imputed_for_model)
sum(hmda_data_pa_df_imputed_for_model$co_applicant_present == 1)
sum(hmda_data_pa_df_imputed_for_model$co_applicant_present == 0)

hmda_data_pa_df_imputed_for_model$applicant_race_and_ethnicity <- as.factor(hmda_data_pa_df_imputed_for_model$applicant_race_and_ethnicity)

hmda_data_pa_df_imputed_for_model$co_applicant_sex <- as.factor(hmda_data_pa_df_imputed_for_model$co_applicant_sex)

hmda_data_pa_df_imputed_for_model$co_applicant_present <- as.factor(hmda_data_pa_df_imputed_for_model$co_applicant_present)

hmda_data_pa_df_imputed_for_model$county_code <- hmda_data_pa_df_imputed_for_model$county_code

hmda_data_pa_df_imputed_for_model$hoepa_status <- as.factor(hmda_data_pa_df_imputed_for_model$hoepa_status)

hmda_data_pa_df_imputed_for_model$lien_status <- as.factor(hmda_data_pa_df_imputed_for_model$lien_status)

hmda_data_pa_df_imputed_for_model$preapproval <- as.factor(hmda_data_pa_df_imputed_for_model$preapproval)

hmda_data_pa_df_imputed_for_model$property_type <- as.factor(hmda_data_pa_df_imputed_for_model$property_type)

hmda_data_pa_df_imputed_for_model$purchaser_type <- as.factor(hmda_data_pa_df_imputed_for_model$purchaser_type)

hmda_data_pa_df_imputed_for_model$applicant_sex <- as.factor(hmda_data_pa_df_imputed_for_model$applicant_sex)

hmda_data_pa_df_imputed_for_model$county_code <- as.factor(hmda_data_pa_df_imputed_for_model$county_code)

hmda_data_pa_df_imputed_for_model$county_name <- as.factor(hmda_data_pa_df_imputed_for_model$county_name)

hmda_data_pa_df_imputed_for_model$agency_code <- as.factor(hmda_data_pa_df_imputed_for_model$agency_code)

hmda_data_pa_df_imputed_for_model$owner_occupancy <- as.factor(hmda_data_pa_df_imputed_for_model$owner_occupancy)

head(hmda_data_pa_df_imputed_for_model)

hmda_data_pa_df_imputed_for_model$log_loan_amount_000s <- log(hmda_data_pa_df_imputed_for_model$loan_amount_000s)

hmda_data_pa_df_imputed_for_model$log_applicant_income_000s <- log(hmda_data_pa_df_imputed_for_model$applicant_income_000s)
str(hmda_data_pa_df_imputed_for_model)
```

# Train test split. Stratified sampling

```{r}
train_test_split <- function(input_df, train_proportion = 0.8) {
  library(caret)
  
  # https://stackoverflow.com/questions/20776887/stratified-splitting-the-data
  train.index <- createDataPartition(y = input_df$loan_granted, p = train_proportion, list = FALSE)
  
  train <- input_df[train.index, ]
  test <-  input_df[-train.index, ]
  
  train$loan_granted <- as.factor(train$loan_granted)
  test$loan_granted <- as.factor(test$loan_granted)
  
  val_list <- list(train, test)
  return (val_list)
}

```

# Generate train/test split.

```{r}
sum(hmda_data_pa_df_imputed_for_model$loan_granted == TRUE)
sum(hmda_data_pa_df_imputed_for_model$loan_granted == FALSE)
nrow(hmda_data_pa_df_imputed)

split_dfs <- train_test_split(hmda_data_pa_df_imputed_for_model[, -which(names(hmda_data_pa_df_imputed_for_model) %in% c("co_applicant_race_and_ethnicity", "county_name"))])
train <- split_dfs[[1]]
test <- split_dfs[[2]]

head(train)
colnames(train)
unique(train$loan_granted)
```


# Analysis on the train and test split datasets.
```{r}

writeLines("")
writeLines("Possible values for applicant_race_and_ethnicity")
levels(hmda_data_pa_df_imputed_for_model$applicant_race_and_ethnicity)

writeLines("")
writeLines("[train df]. Race and ethnicity numbers for whites and African Americans")
nrow(train)
sum(train$applicant_race_and_ethnicity == "White")
sum(train$applicant_race_and_ethnicity == "Black or African American")

writeLines("")
writeLines("Generating upsampled training dataframe based on race and ethnicity")
set.seed(1000)

up_train <- upSample(x = train, y = train$applicant_race_and_ethnicity)

writeLines("")
writeLines("[Upsampled train df]. Race and ethnicity numbers for whites and African Americans")

sum(as.character(up_train$applicant_race_and_ethnicity) == "White")
sum(as.character(up_train$applicant_race_and_ethnicity) == "Black or African American")
nrow(up_train)

writeLines("")
writeLines("[test df].Race and ethnicity numbers for whites and African Americans")

nrow(test)
sum(test$applicant_race_and_ethnicity == "White")
sum(test$applicant_race_and_ethnicity == "Black or African American")

```

# Models

# Model 1 without loan to income ratio.
```{r}
# Initial model. Predictors selected are based on intuition.
# Please update this list as needed.
#model <- glm(loan_granted~applicant_race_and_ethnicity+applicant_income_000s+co_applicant_race_and_ethnicity+co_applicant_sex+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+co_applicant_present, family = binomial(link='logit'), data=train)

model <- glm(loan_granted~applicant_race_and_ethnicity+applicant_income_000s+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+co_applicant_present+agency_code+minority_population, family = binomial(link='logit'), data=train)


writeLines("Summary for model is as below")
writeLines("")
summary(model)

writeLines("Anova for model is as below")
writeLines("")
anova(model)
```

# Model plots
```{r}
plot(model)
```

# Model with log of loan amount and income.
```{r}
model_log_income_and_amount <- glm(loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+log_loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+co_applicant_present+agency_code+minority_population, family = binomial(link='logit'), data=train)

writeLines("Summary for log model is as below")
writeLines("")
summary(model_log_income_and_amount)

writeLines("Anova for log model is as below")
writeLines("")
anova(model_log_income_and_amount)

```

# Model 2 with loan to income ratio.

```{r}
model_loan_income_ratio <- glm(loan_granted~applicant_race_and_ethnicity+applicant_income_000s+loan_to_income_ratio+co_applicant_present+co_applicant_sex+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code+minority_population, family = binomial(link='logit'), data=train)

writeLines("Summary for model is as below")
writeLines("")
summary(model_loan_income_ratio)

writeLines("Anova for model is as below")
writeLines("")
anova(model_loan_income_ratio)

```

# Model plots
```{r}
plot(model_loan_income_ratio)
```

# Model 2 with log of loan amounts and income and loan to income ratio.

```{r}
model_log_amount_loan_income_ratio <- glm(loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+loan_to_income_ratio+co_applicant_present+co_applicant_sex+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+log_loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code+minority_population, family = binomial(link='logit'), data=train)

writeLines("Summary for log model with loan to income ratio is as below")
writeLines("")
summary(model_log_amount_loan_income_ratio)

writeLines("Anova for model is as below")
writeLines("")
anova(model_log_amount_loan_income_ratio)

```

# Prediction based on the base model.

```{r}
# TODO
# Look into the rank-deficient fit warning here.
# Seems related to correlated predictors.
fitted.probs <- predict(model, test, type = "response")
```

# Compute statistics like classification errror, confusion matrix, etc based
# on our predictions.

```{r}
prediction_stats <- function(fitted.probs, test, threshold = 0.5) {
  fitted.results <- ifelse(fitted.probs > threshold, TRUE, FALSE)
  
  classification_error <- mean(fitted.results != test$loan_granted)
  classification_error
  
  print(paste('Accuracy', 1 - classification_error))
  
  print(confusionMatrix(as.factor(fitted.results), as.factor(test$loan_granted), positive = "TRUE"))
}
```
## Function to calculate precision and recall using confusion matrix
```{r}
precision_recall <- function(confusion_matrix){
  precision <- confusion_matrix[4] / (confusion_matrix[4]+confusion_matrix[2])
  print("Precision")
  print(precision)
  
  recall <- confusion_matrix[4] / (confusion_matrix[4]+confusion_matrix[3])
  print("Recall")
  print(recall)
}
```

# Model performance.
```{r}
performance <- prediction_stats(fitted.probs, test)
performance
```
## Precision and Recall
```{r}
precision_recall(performance$table)
```

# Plot TPR vs FPR
```{r}
library(ROCR)

#https://www.r-bloggers.com/evaluating-logistic-regression-models/#:~:text=K%2DFold%20Cross%20Validation&text=One%20such%20technique%20for%20doing,variable%20in%20our%20testing%20data.
pred <- prediction(fitted.probs, test$loan_granted)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf)

```

# Prediction based on the model with loan to income ratio
```{r}
# TODO
# Look into the rank-deficient fit warning here.
# Seems related to correlated predictors.
fitted_2.probs <- predict(model_loan_income_ratio, test, type = "response")
```

# Model performance.
```{r}
performance<-prediction_stats(fitted_2.probs, test)
performance
```
## Precision and Recall
```{r}
precision_recall(performance$table)
```

# Prediction based on the model with log income and loan to income ratio
```{r}
# TODO
# Look into the rank-deficient fit warning here.
# Seems related to correlated predictors.

fitted_2_log_income.probs <- predict(model_log_amount_loan_income_ratio, test, type = "response")
```

# Model performance.
```{r}
performance<-prediction_stats(fitted_2_log_income.probs, test)
performance
```
```{r}
precision_recall(performance$table)
```

# Trying out cross validation to evaluate the performance.
```{r}
ctrl <- trainControl(method = "cv", number = 10, savePredictions = TRUE)

model_base_cross_val <- train(loan_granted~applicant_race_and_ethnicity+applicant_income_000s+co_applicant_present+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+purchaser_type+loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code+minority_population, data=train, method="glm", family="binomial", trControl = ctrl, tuneLength = 15)

```

# Trying out cross validation with log income to evaluate the performance.
```{r}
ctrl <- trainControl(method = "cv", number = 10, savePredictions = TRUE)
model_base_cross_val_log_amounts <- train(loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+co_applicant_present+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+purchaser_type+log_loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code+minority_population, data=train, method="glm", family="binomial", trControl = ctrl, tuneLength = 15)

```

# Cross validation including loan to income ratio
```{r}
library(caret)
ctrl <- trainControl(method = "cv", number = 10, savePredictions = TRUE)

model_cross_val_loan_income_ratio <- train(loan_granted~co_applicant_present+co_applicant_sex+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+purchaser_type+loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code+minority_population, data=train, method="glm", family="binomial", trControl = ctrl, tuneLength = 5)

```

# Cross validation including loan to income ratio
```{r}
library(caret)
ctrl <- trainControl(method = "cv", number = 10, savePredictions = TRUE)

model_cross_val_loan_income_ratio_log_amount <- train(loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+loan_to_income_ratio+co_applicant_present+co_applicant_sex+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+purchaser_type+log_loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code+minority_population, data=train, method="glm", family="binomial", trControl = ctrl, tuneLength = 5)

```
# Predictions on the base model computed from k-fold cross validation
```{r}
pred.probs = predict(model_base_cross_val, newdata=test, type = "prob")
pred.results <- ifelse(pred.probs[,2] > 0.5, TRUE, FALSE)
performance = confusionMatrix(data=as.factor(pred.results), test$loan_granted, positive = "TRUE")
precision_recall(performance$table)
```

# Predictions on the model with loan income ratio computed from k-fold cross validation
```{r}
pred2.probs = predict(model_cross_val_loan_income_ratio, newdata=test, type = "prob")
pred2.results <- ifelse(pred2.probs[,2] > 0.5, TRUE, FALSE)
performance<-confusionMatrix(data=as.factor(pred2.results), test$loan_granted, positive = "TRUE")
performance
precision_recall(performance$table)
```

# Predictions on the base model with log amounts computed from k-fold cross validation
```{r}
pred_log_k_fold.probs = predict(model_base_cross_val_log_amounts, newdata=test, type = "prob")
pred.results <- ifelse(pred_log_k_fold.probs[,2] > 0.5, TRUE, FALSE)
performance<-confusionMatrix(data=as.factor(pred.results), test$loan_granted, positive = "TRUE")
performance
precision_recall(performance$table)
```

# Predictions on the model with log amounts and loan income ratio computed from k-fold cross validation
```{r}
pred2_log_k_fold_loan_income_ratio.probs = predict(model_cross_val_loan_income_ratio_log_amount, newdata=test, type = "prob")
pred2_log_k_fold_loan_income_ratio.results <- ifelse(pred2_log_k_fold_loan_income_ratio.probs[,2] > 0.5, TRUE, FALSE)
performance<-confusionMatrix(data=as.factor(pred2_log_k_fold_loan_income_ratio.results), test$loan_granted, positive = "TRUE")
performance
precision_recall(performance$table)

```
# Save the imputed dataframe to file.
```{r}
write.csv(hmda_data_pa_df_imputed, "D:/data/HMDA/2014/hmda_2014_pa_imputed.csv", row.names = FALSE)
```

# Model with decision tree.
```{r}
library(rpart)

model_dt_loan_to_income_ratio <- rpart(loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+loan_to_income_ratio+co_applicant_present+co_applicant_sex+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+log_loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code+minority_population, data=train, method = 'class')
```

```{r}
library(rpart.plot)
rpart.plot(model_dt_loan_to_income_ratio, extra = 106)

```



# Predictions using dt.

```{r}
preds_dt <- predict(model_dt_loan_to_income_ratio, test, type = 'class')

```

```{r}
confusionMatrix(data=as.factor(preds_dt), test$loan_granted, positive = "TRUE")
```

# R logistic regression stepAIC steps

```{r}
full_model <- glm(train$loan_granted~., data = train, family = binomial(link='logit'))
summary(full_model)
```


```{r}
library(MASS)
aic_model <- full_model %>% stepAIC(trace = TRUE)
```

# Comparing full and stepAIC models

```{r}

writeLines("")
writeLines("Summary of full model is as below")
writeLines("")
summary(full_model)

writeLines("")
writeLines("Summary of stepAIC model are as below")
writeLines("")
summary(aic_model)

```

# Compare predictions with full and AIC models

```{r}

unique(test$purchaser_type)
unique(train$purchaser_type)

writeLines("")
writeLines("Prediction stats from full model")
writeLines("")
full_model.probs <- predict(full_model, test, type = "response")
prediction_stats(full_model.probs, test)

writeLines("")
writeLines("Prediction stats from stepAIC model")
writeLines("")
aic_model.probs <- predict(aic_model, test, type = "response")
prediction_stats(aic_model.probs, test)

```

# load the census dataset containing poplation distribution per county.
```{r}
library(dplyr)
census_data_dir <- "d:/data/census"

census_pa_file_path <- paste(census_data_dir, sep = "", "/cc-est2018-alldata-42.csv")

census_pa_data <- fread(census_pa_file_path)

census_pa_data_df <- as.data.frame(census_pa_data)

census_pa_data_df <- filter(census_pa_data_df, census_pa_data_df$AGEGRP == 0 & census_pa_data_df$YEAR == 1)

census_pa_data_df$WHITE_PERCENT <- ((census_pa_data_df$WA_MALE + census_pa_data_df$WA_FEMALE) / census_pa_data_df$TOT_POP) * 100

census_pa_data_df$BLACK_PERCENT <- ((census_pa_data_df$BA_MALE + census_pa_data_df$BA_FEMALE) / census_pa_data_df$TOT_POP) * 100

census_pa_data_df$ASIAN_PERCENT <- ((census_pa_data_df$AA_MALE + census_pa_data_df$AA_FEMALE) / census_pa_data_df$TOT_POP) * 100

census_pa_data_df$NATIVE_HAWAII_PACIFIC_ISLANDER_PERCENT <- ((census_pa_data_df$NA_FEMALE + census_pa_data_df$NA_MALE) / census_pa_data_df$TOT_POP) * 100

census_pa_data_df$AMERICAN_INDIAN_PERCENT <- ((census_pa_data_df$IA_FEMALE + census_pa_data_df$IA_FEMALE) / census_pa_data_df$TOT_POP) * 100

census_pa_data_df$HISPANIC_PERCENT <- ((census_pa_data_df$H_MALE + census_pa_data_df$H_FEMALE) / census_pa_data_df$TOT_POP) * 100

census_pa_data_df$CTYNAME <- as.factor(census_pa_data_df$CTYNAME)
class(census_pa_data_df$CTYNAME)

```

# Visualize the census data frame.
```{r}

head(census_pa_data_df)

str(census_pa_data_df)

```


```{r}
table1 <- as.data.table(hmda_data_pa_df_imputed_for_model)
table2 <- as.data.table(census_pa_data_df[, c("WHITE_PERCENT", "BLACK_PERCENT", "ASIAN_PERCENT", "NATIVE_HAWAII_PACIFIC_ISLANDER_PERCENT", "AMERICAN_INDIAN_PERCENT", "HISPANIC_PERCENT", "CTYNAME")])

join <- merge(table1, table2, by.x = c("county_name"), by.y = c("CTYNAME"))

merged_pa_df_imputed_for_model <- as.data.frame(join)

merged_pa_df_imputed_for_model$log_loan_amount_000s <- log(merged_pa_df_imputed_for_model$loan_amount_000s)

merged_pa_df_imputed_for_model$log_applicant_income_000s <- log(merged_pa_df_imputed_for_model$applicant_income_000s)


```

# Train test split for the merged data frame created above
```{r}

merged_split_dfs <- train_test_split(merged_pa_df_imputed_for_model[, -which(names(merged_pa_df_imputed_for_model) %in% c("co_applicant_race_and_ethnicity", "county_name"))])

merged_train <- merged_split_dfs[[1]]
merged_test <- merged_split_dfs[[2]]

head(merged_train)

```


# Build model using the merged training set computed above.
```{r}
# Removing county code as it is correlated with the BLACK_PERCENT, WHITE_PERCENT
# variables computed above. Leads to singularity warnings.
merged_model <- glm(loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+log_loan_amount_000s+msamd+applicant_sex+hud_median_family_income+tract_to_msamd_income+co_applicant_present+agency_code+BLACK_PERCENT+ASIAN_PERCENT+HISPANIC_PERCENT+loan_to_income_ratio, family = binomial(link='logit'), data=merged_train)

```


# Analysis on the model created above
```{r}
writeLines("Summary for merged model is as below")
writeLines("")
summary(merged_model)

writeLines("Anova for merged model is as below")
writeLines("")
anova(merged_model)

```

# Prediction based on the merged model.

```{r}
# TODO
# Look into the rank-deficient fit warning here.
# Seems related to correlated predictors.
fitted_merged.probs <- predict(merged_model, merged_test, type = "response")
```

# Merged model performance.
```{r}
prediction_stats(fitted_merged.probs, merged_test)
```


# Cross validation on the merged data frame including loan to income ratio
```{r}
library(caret)
ctrl <- trainControl(method = "cv", number = 10, savePredictions = TRUE)

merged_model_cross_val <- train(loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+log_loan_amount_000s+msamd+applicant_sex+hud_median_family_income+tract_to_msamd_income+co_applicant_present+agency_code+BLACK_PERCENT+ASIAN_PERCENT+HISPANIC_PERCENT+loan_to_income_ratio, data=merged_train, method="glm", family="binomial", trControl = ctrl, tuneLength = 5)

```

# Predictions on the merged model with loan income ratio computed from k-fold cross validation
```{r}
pred_merged.probs = predict(merged_model_cross_val, newdata=merged_test, type = "prob")
pred_merged.results <- ifelse(pred_merged.probs[,2] > 0.5, TRUE, FALSE)
performance = confusionMatrix(data=as.factor(pred_merged.results), merged_test$loan_granted, positive = "TRUE")
precision_recall(performance$table)
```
