---
title: "HMDA_2008_IL_EDA - Initial Analysis for 2008 IL Data"
author: "Virat Joshi"
date: "4/5/2020"
output: html_document
---

### Set working directory
```{r}
working_dir = "/Users/vjstark/Downloads/CSP-571/Project"
setwd(working_dir)

data_dir <- "./Data"
```

## Load essential Libraries
```{r}
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(corrplot)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)
library(e1071)
library(caret)
```
## Load data file
```{r}
hmda_data_pa <- fread(paste(data_dir, "/hmda_2008_pa_all-records_labels.csv", sep = ""))
```
## Convert to dataframe
```{r}
hmda_pa_df <- as.data.frame(hmda_data_pa)
head(hmda_pa_df)
```


# Data Analysis Section 1 : Understanding the data
### Columns in the dataset
```{r}
colnames(hmda_pa_df)
```

### Dimensions of the data
```{r}
dim(hmda_pa_df)
```

### First 10 rows from the data
```{r}
head(hmda_pa_df,10)
```

### Print a glimpse of the dataset
```{r}
glimpse(hmda_pa_df)
```

# Data Analysis Section 2: Missing Values
We start by finding the number of NAs in the dataset
```{r}
sapply(hmda_pa_df,function(x) sum(is.na(x)))
```

Checking the dattaframe for empty strings
```{r}
sapply(hmda_pa_df,function(x) sum(x==""))
```

Check for missing values present as ?
```{r}
sapply(hmda_pa_df,function(x) sum(x=="?"))
```

Check for NULL values
```{r}
sapply(hmda_pa_df,function(x) sum(is.null(x)))
```

# Data Analysis Section 3: Analysis on the columns of interest
### Columns having the information about applicants Race and Ethnicity, 
Applicants Ethnicity Values
```{r}
unique(hmda_pa_df$applicant_ethnicity_name)
```
Applicant Race Encoded Numbers
```{r}
unique(hmda_pa_df$applicant_race_1)
```

Applicant Race Name
```{r}
unique(hmda_pa_df$applicant_race_name_1)
```

Distribution of applicants based on their race
```{r}
hmda_pa_df$applicant_race_and_ethnicity <- NA
hmda_pa_df$co_applicant_race_and_ethnicity <- NA

hmda_pa_df$applicant_race_and_ethnicity <- ifelse(hmda_pa_df$applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_pa_df$applicant_race_name_1)

hmda_pa_df$co_applicant_race_and_ethnicity <- ifelse(hmda_pa_df$co_applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_pa_df$co_applicant_race_name_1)

hmda_pa_df$applicant_race_and_ethnicity[hmda_pa_df$applicant_race_and_ethnicity == "Information not provided by applicant in mail, Internet, or telephone application"] <- "No Information Provided"

head(hmda_pa_df)
```

```{r}
count_racename <- as.data.frame(table(hmda_pa_df$applicant_race_and_ethnicity))
colnames(count_racename)<-c("Race_and_Ethnicity","Count")
p<-ggplot(data=count_racename, aes(x=reorder(Race_and_Ethnicity,-Count), y=Count)) + geom_bar(stat="identity",fill="steelblue")+geom_text(aes(label=Count), size=2.5)
p + coord_flip() +labs(title = "Distribution of applicants by Race and Ethnicity", y="Number of Applicants",x="Race and Ethnicity")

```
The number of white applicants are comparatively higher in the dataset
### Applicants Income

Lets dive into the applicants income
```{r}
summary(hmda_pa_df$applicant_income_000s)
```
Here it can be observed that the income of applicants is spreadout a lot. We have applicants with income ranging from 1 to 9999 in thousands. Also, missing values sum upto 42959.

### Boxplot for the people having income less that 500K
```{r}
temp<-as.data.frame(hmda_pa_df$applicant_income_000s[hmda_pa_df$applicant_income_000s<500])
ggplot(data = temp, aes(x=hmda_pa_df$applicant_income_000s[hmda_pa_df$applicant_income_000s < 500])) +geom_histogram(fill = "steelblue") + labs(title = "Applicant for income below 500K" , x = "Income in thousands",y="Count")
```

The distribution is right skewed. It would be better if we use median as the measure of central tendency. Lets see what % of applicants have income below 500 from the entire dataset.
```{r}
no_of_NAs <- sum(is.na(temp))
income_below_500<-length(hmda_pa_df$applicant_income_000s[hmda_pa_df$applicant_income_000s<500])
total_applicants<-length(hmda_pa_df$applicant_income_000s)

((income_below_500 - no_of_NAs ) / total_applicants ) *100
```
Approximately 93.49% of the total applicants have income below 500K.

### Applicants according to their race and there median income.

```{r}
m <- summarise_at(group_by(hmda_pa_df,applicant_race_and_ethnicity),vars(applicant_income_000s),funs(median(.,na.rm=TRUE)))
ggplot(data = m ,aes (x =  applicant_race_and_ethnicity,y = applicant_income_000s)) + geom_bar(stat = "identity",fill = "steelblue") +geom_text(aes(label=applicant_income_000s), size=3,nudge_y = 5)+ coord_flip()+labs(title = "Median of applicants income by Race and Ethnicity", y="Median Income",x="Race and Ethnicity")
```
The mean income for Asians is the hgihest followed by applicants with no information about race looks high. 
This maybe due to the following reason.

```{r}
p <- ggplot(group_by(hmda_pa_df,applicant_race_and_ethnicity), aes(x=applicant_race_and_ethnicity, y=applicant_income_000s)) + 
  geom_boxplot() + labs(title = "Applicants Income According to Race And Ethnicity" , x = "Applicant Race and Ethnicity" , y = "Income distribution in thousands")
p +coord_flip()
```
In the boxplot, the category where people didn’t provide the race and ethnicity data has most extreme values of income. These are probably outliers. This is why the mean of certain categories shifted to the higher end. Lets look in this data

### Distribution of income for the originated loans
```{r}
hmda_origination_status_df <- hmda_pa_df[hmda_pa_df$action_taken == "1", ]
summary(hmda_origination_status_df$applicant_income_000s)
```

```{r}
p <- ggplot(group_by(hmda_origination_status_df,applicant_race_and_ethnicity), aes(x=applicant_race_and_ethnicity, y=applicant_income_000s)) + 
  geom_boxplot() + labs(title = "Approved Loan Applicant's Income Distribution" , x = "Applicant Race and Ethnicity" , y = "Income distribution in thousands")
p +coord_flip()
```

```{r}
m <- summarise_at(group_by(hmda_origination_status_df,applicant_race_and_ethnicity),vars(applicant_income_000s),funs(median(.,na.rm=TRUE)))
ggplot(data = m ,aes (x =  applicant_race_and_ethnicity,y = applicant_income_000s)) + geom_bar(stat = "identity",fill = "steelblue") +geom_text(aes(label=applicant_income_000s), size=3 , nudge_y = 5)+ coord_flip()+labs(title = "Median of applicants income by Race and Ethnicity", y="Median Income",x="Race and Ethnicity")

```

```{r}
temp<-as.data.frame(hmda_origination_status_df$applicant_income_000s[hmda_origination_status_df$applicant_income_000s<500])

ggplot(data = temp, aes(x=hmda_origination_status_df$applicant_income_000s[hmda_origination_status_df$applicant_income_000s < 500])) +geom_histogram(fill = "steelblue") + labs(title = "Applicant with loan approved and income below 500K" , x = "Income in thousands",y="Count")
```

### Proportion of loans originated 
What are the set of decisions made for an application?
```{r}
unique(hmda_pa_df$action_taken_name)
```
Here, we can say the loan is denied in two ways. 
1. Application denied by financial institution
2. Preapproval request denied by financial institution

The loan is approved if 
1. "Application approved but not accepted"
2. "Loan originated"

Other than this, we would classify decision as "Technical or paperwork Issues"
1. "Application withdrawn by applicant"
2. "Loan purchased by the institution
3. "File closed for incompleteness"
4. "Preapproval request approved but not accepted"

```{r}
hmda_pa_df$Decision <- NA

hmda_pa_df$Decision <- ifelse(hmda_pa_df$action_taken_name == "Application denied by financial institution" | hmda_pa_df$action_taken_name == "Preapproval request denied by financial institution" , "Denied" ,  ifelse(hmda_pa_df$action_taken_name == "Loan originated" | hmda_pa_df$action_taken_name == "Application approved but not accepted" ,  "Approved" , "Paperwork Issues"))
      
tbl <- with(hmda_pa_df, table(Decision,applicant_race_and_ethnicity))

ggplot(as.data.frame(tbl), aes(factor(applicant_race_and_ethnicity), Freq, fill = Decision)) +     
  geom_col(position = 'dodge') + labs(title = "Proportion of loans denied by financial Institutions",x = "Race and Ethnicity" , y = "Number of applicants") + coord_flip()
```

```{r}
tbl <- with(hmda_pa_df, table(action_taken_name,applicant_race_and_ethnicity))

ggplot(as.data.frame(tbl), aes(factor(applicant_race_and_ethnicity), Freq, fill = action_taken_name)) +     
  geom_col(position = 'dodge') + labs(title = "Proportion of loans denied by financial Institutions",x = "Race and Ethnicity" , y = "Number of applicants") 
```

```{r}
mortgage_status_by_race_and_ethnicity <- hmda_pa_df %>% group_by(action_taken_name, applicant_race_and_ethnicity) %>%
  summarise(ActionCount = n()) %>%
  arrange(desc(ActionCount))

mortgage_by_race_and_ethnicity = hmda_pa_df %>% group_by(applicant_race_and_ethnicity) %>%
  summarise(EthnicityCount = n()) %>%
  arrange(desc(EthnicityCount))

mortgage_status_aggregated_by_race_and_ethnicity  = inner_join(mortgage_status_by_race_and_ethnicity, mortgage_by_race_and_ethnicity) %>% mutate(percentage = (ActionCount / EthnicityCount) * 100)

ggplot(mortgage_status_aggregated_by_race_and_ethnicity, aes(x = reorder(action_taken_name, percentage), 
                                          y = percentage)) +
  geom_bar(stat='identity', fill ="#00AFBB") +
  facet_wrap(~ applicant_race_and_ethnicity) +
  geom_text(aes(x = action_taken_name, y = 1, label = paste0("(",round(percentage),"%)",sep="")),
            hjust=0, vjust=.5, size = 2.5, colour = 'black',
            fontface = 'bold') +
  labs(x = 'Action', y = 'Percent Count Of Action Taken', title = 'Actions for Loan Application by Race') +
  coord_flip() + 
  theme_bw()
```

Some colors which can be used for graphs
```{r}

cbp2 <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```


### Different Type of Loans
```{r}
unique(hmda_pa_df$loan_type_name)
```


```{r}
ggplot(data = summarise_at(group_by(hmda_pa_df,loan_type_name),vars(loan_type),funs(n())),aes(x = loan_type_name,y =  loan_type)) + geom_bar(stat = "identity",fill = "#009E73") + geom_text(aes(label = loan_type), vjust = -0.5) +labs(title = "Type of Loans Distribution" , x = "Loan Type" , y = "Count")
```

### Applicants  loan amount
```{r}
skew <- paste("Skewness:",skewness(hmda_pa_df$loan_amount_000s,na.rm = TRUE))
ggplot(data = hmda_pa_df , aes(x = loan_amount_000s)) + geom_histogram(fill = "steelblue") + labs(title = "Loan amount distribution" , x = "Loan amount in thousands" , y = "Count")+ annotate("text", x = 100000, y = 300000, size = 3.2,label = skew)
```

Looks like the data is highly skewed.
```{r}
#install.packages("moments")
library(moments)
skewness(hmda_pa_df$loan_amount_000s,na.rm = TRUE)
```
The data for loan amount is highly right skewed. Changes should be made so that the prediction model does not mess up.

## Handling highly skewed data. Log Transformation
```{r}
skew <- paste("Skewness:",skewness(log(hmda_pa_df$loan_amount_000s),na.rm = TRUE))
ggplot(data = hmda_pa_df , aes(x = log(loan_amount_000s))) + geom_histogram(fill = "steelblue") + labs(title = "Log transformed distribution for Loan amount" , x = "log(Loan Amount)", y = 'Count')+ annotate("text", x = 8, y = 100000, size = 3.2,label = skew)
```

```{r}
skewness(log(hmda_pa_df$loan_amount_000s),na.rm = TRUE)
```
After transforming the values, the data is slightly left skewed but the value is better than without transformation.
### Same is the case with applicants income
```{r}
skew <- paste("Skewness:",skewness(hmda_pa_df$applicant_income_000s,na.rm = TRUE))
ggplot(data = hmda_pa_df , aes(x = applicant_income_000s)) + geom_histogram(fill = "steelblue") + labs(title = "Applicant Income distribution" , x = "Applicant Income in thousands" , y = "Count") + annotate("text", x = 100000, y = 90000, size = 3.2,label = skew)
```

```{r}
skew <- paste("Skewness:",skewness(log(hmda_pa_df$applicant_income_000s),na.rm=TRUE))
ggplot(data = hmda_pa_df , aes(x = log(applicant_income_000s))) + geom_histogram(fill = "steelblue") + labs(title = "Log transformed distribution for Applicant Income" , x = "log(Applicant Income)", y = 'Count') +annotate("text", x = 10, y = 90000, size = 3.2,label = skew)
```


### Which columns have most values missing?

```{r}
hmda_pa_df<-data.frame(hmda_data_pa)

hmda_pa_df$applicant_race_and_ethnicity <- NA
hmda_pa_df$co_applicant_race_and_ethnicity <- NA

hmda_pa_df$applicant_race_and_ethnicity <- ifelse(hmda_pa_df$applicant_ethnicity_name == "Hispanic or Latino",
                                                    "Hispanic or Latino", hmda_pa_df$applicant_race_name_1)

hmda_pa_df$co_applicant_race_and_ethnicity <- ifelse(hmda_pa_df$co_applicant_ethnicity_name == "Hispanic or Latino",
                                                       "Hispanic or Latino", hmda_pa_df$co_applicant_race_name_1)

m<-sapply(hmda_pa_df, function(x) sum(is.na(x)))
temp<-c()
for (i in 1:length(m)){
  if(m[i]>0){
    temp<-c(temp,m[i])
  }
}
missing <- data.frame(temp)
missing$percent <- (missing$temp / length(hmda_pa_df$as_of_year))*100
missing
```

From the dataframe above, we can see that there are a lot of columns with missing values more that 99%. We can't impute these values with very less prior data. Also, these columns would not provide any information while making the decision of whether to grant a loan or not. 

Other perpective to look at this is, a lot of applicants don't have the information about these predictors. So most probably, banks don't use this information to make decision.

Hence, we can get rid of this columns. This can reduce dimensionality of our dataset.


```{r}
drop<- c ("applicant_race_2",
          "applicant_race_3",
          "applicant_race_4",
          "applicant_race_5",
          "co_applicant_race_2",
          "co_applicant_race_3",
          "co_applicant_race_4",
          "co_applicant_race_5",
          "denial_reason_1",
          "denial_reason_2",
          "denial_reason_3",
          "edit_status_name",
          "edit_status",
          "rate_spread",
          "sequence_number",
          "application_date_indicator",
          "as_of_year", # Does not provide extra information which can be used
          "respondent_id", # Does not provide extra information which can be used
          "agency_abbr", # Does not provide extra information which can be used
          "agency_name", # Does not provide extra information which can be used
          "loan_type_name", # We are only dealing with conventional loans. 
          "loan_type",
          "state_abbr",# Not needed as data is of same state
          "state_code", # Not needed as data is of same state
          "owner_occupancy_name", # Redundant column. Encoded values present
          "msamd_name", # Redundant column. Encoded values present
          "state_name", # Redundant column. Encoded values present
          "applicant_sex_name", # Redundant column. Encoded values present
          "hoepa_status_name", # Redundant column. Encoded values present
          "lien_status_name", # Redundant column. Encoded values present
          "county_name", # Redundant column. Encoded values present
          "preapproval_name", # Redundant column. Encoded values present
          "property_type_name", # Redundant column. Encoded values present
          "action_taken_name", # Redundant column. Encoded values present
          "loan_purpose_name", # Redundant column. Encoded values present
          "purchaser_type_name", # Redundant column. Encoded values present
          "applicant_ethnicity_name", # Redundant column. Encoded values present
          "co_applicant_ethnicity_name", # Redundant column. Encoded values present
          "co_applicant_sex_name" ) # Redundant column. Encoded values present
hmda_pa_df<-hmda_pa_df[ , !(names(hmda_pa_df) %in% drop)]
dim(hmda_pa_df)
```

```{r}
hmda_pa_df$loan_granted <- ifelse(hmda_pa_df$action_taken == 1 | hmda_pa_df$action_taken == 6, 1, 0)
```

```{r}
colnames(hmda_pa_df)
```

Make the labels binary. Currently the action_taken variable is multivariate. We classify the loan is originated when it has "Loan Originated" or ""Loan purchased by the institution" label. Then drop this column as it is corelated to the new column we generate. 

```{r}
hmda_pa_df$loan_granted <- ifelse(hmda_pa_df$action_taken == 1 | hmda_pa_df$action_taken == 6, TRUE, FALSE)
d <-c("action_taken")
hmda_pa_df<-hmda_pa_df[ , !(names(hmda_pa_df) %in% d)]
```

### As we have narrowed the list down, lets look which factors influence the decision whether the loan is granted or not.
For this we can use XGBoost to find the feature importances. This can help us get started and use the predictors which influences the decision. 

```{r}
dmy <- dummyVars(" ~ .", data = hmda_pa_df, fullRank=T)
df <- data.frame(predict(dmy, newdata = hmda_pa_df))

df = df %>% select(-loan_grantedTRUE)
df$loan_granted = hmda_pa_df$loan_granted

features<-colnames(df)

for (f in features) {
  if ((class(df[[f]])=="factor") || (class(df[[f]])=="character")) {
    levels <- unique(df[[f]])
    df[[f]] <- as.numeric(factor(df[[f]], levels=levels))
  }
}
```


```{r}
formula = loan_granted ~ .

fitControl <- trainControl(method="none",classProbs = TRUE)

xgbGrid <- expand.grid(nrounds = 100,
                       max_depth = 3,
                       eta = .05,
                       gamma = 0,
                       colsample_bytree = .8,
                       min_child_weight = 1,
                       subsample = 1)


set.seed(13)
```



```{r}
#install.packages()
df$loan_granted = as.factor(df$loan_granted)
levels(df$loan_granted) = c("No","Yes")

XGB = train(formula, data = df,
                   method = "xgbTree",trControl = fitControl,
                   tuneGrid = xgbGrid,na.action = na.pass)


importance = varImp(XGB)



varImportance <- data.frame(Variables = row.names(importance[[1]]), 
                            Importance = round(importance[[1]]$Overall,2))

rankImportance <- varImportance %>%
  mutate(Rank = paste0('#',dense_rank(desc(Importance)))) %>%
  head(30)

rankImportancefull = rankImportance
print(rankImportance)
ggplot(rankImportance, aes(x = reorder(Variables, Importance), 
                           y = Importance)) +
  geom_bar(stat='identity',colour="white", fill = "steelblue") +
  geom_text(aes(x = Variables, y = 1, label = Rank),
            hjust=0, vjust=.5, size = 4, colour = 'black',
            fontface = 'bold') +
  labs(x = 'Variables', title = 'Relative Variable Importance') +
  coord_flip() + 
  theme_bw()
```



