 ---
title: "CSP571_Project_Experimentation for Pennsylvania conventional home loans for 2015"
authors: "Anantanarayanan G Iyengar, [Please add your names here]"
date: "2/17/2020"
output: html_notebook
---

## Global setup like working directory, data directory etc should happen here.
```{r}
# Please change the working directory as needed.
working_dir <- "D:/illinois/csp571/csp-571-02-final-project"
setwd(working_dir)

data_dir <- "D:/data/HMDA"
```

## Install required packages.
```{r}
# https://stackoverflow.com/questions/4090169/elegant-way-to-check-for-missing-packages-and-install-them
list_of_packages <- c("mlbench", "corrplot", "rvest", "tidyr", "stringr", "dplyr", "lubridate", "data.table", "mice", "naniar")
new.packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]

if (length(new.packages)) {
  print("Installing packages\n")
  install.packages(new.packages())
}

library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(naniar)
library(rstudioapi)    

source(paste(dirname(dirname(rstudioapi::getActiveDocumentContext()$path)), "utils/utils.r", sep="/"))

```

## Load data file.
```{r}
hmda_data_pa <- fread(paste(data_dir, "/2015/hmda_2015_pa_all-records_labels.csv", sep = ""))
```

## Data analysis section 1. Print column names and first 10 rows of dataset.
```{r}
hmda_data_pa_df <- as.data.frame(hmda_data_pa)

hmda_data_pa_df <- hmda_data_pa_df[hmda_data_pa_df$loan_type == "1", ]

colnames(hmda_data_pa_df)

writeLines("")

summary(hmda_data_pa_df)

head(hmda_data_pa_df, 10)
```

## Data analysis section 2. Print glimpse of dataset i.e a vertical preview of the dataset.
```{r}
dim(hmda_data_pa_df)
writeLines("Glimpse of hmda dataset for PA")
glimpse(hmda_data_pa_df)

```

## Data analysis section 2. Print structure of dataset.

```{r}
str(hmda_data_pa_df)

```

## Data Analysis Section 3. Check for missing values in the dataset
```{r}
writeLines("Checking for missing values with NA")
sapply(hmda_data_pa_df, function(x) sum(is.na(x)))

writeLines("Checking for missing values with empty strings")
sapply(hmda_data_pa_df, function(x) sum(x == ""))

```

## Look at values in the ethnicity and race columns

```{r}
library(janitor)
writeLines("Application ethnicity values")
levels(as.factor(hmda_data_pa_df$applicant_ethnicity_name))

writeLines("Application race name 1 values")
levels(as.factor(hmda_data_pa_df$applicant_race_1))
levels(as.factor(hmda_data_pa_df$applicant_race_name_1))

hmda_data_pa_df %>% filter(applicant_ethnicity_name == "Hispanic or Latino") %>%
  group_by(applicant_race_name_1) %>% 
           count() %>%
           ungroup() %>%
           replace(is.na(.), 0) %>% 
           adorn_totals(c("col")) %>% 
           arrange(-Total)
```
## Add a new column applicant_race_and_ethnicity and group all applicants
## with ethnicity as Hispanic or Latino as Hispanic or Latino in this column.
## For everyone else, this column gets values from the applicant_race_name_1 
## column

```{r}
hmda_data_pa_df$applicant_race_and_ethnicity <- NA
hmda_data_pa_df$co_applicant_race_and_ethnicity <- NA

hmda_data_pa_df$applicant_race_and_ethnicity <- ifelse(hmda_data_pa_df$applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_data_pa_df$applicant_race_name_1)

hmda_data_pa_df$co_applicant_race_and_ethnicity <- ifelse(hmda_data_pa_df$co_applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_data_pa_df$co_applicant_race_name_1)

#hmda_data_pa_df$applicant_race_and_ethnicity <- hmda_data_pa_df$applicant_race_and_ethnicity

#hmda_data_pa_df$co_applicant_race_and_ethnicity <- hmda_data_pa_df$co_applicant_race_and_ethnicity

head(hmda_data_pa_df)

unique(hmda_data_pa_df$applicant_race_and_ethnicity)
```

## Graph mortgage distribution by applicant race and ethinicity.
```{r}
mortgage_by_race_and_ethnicity = hmda_data_pa_df %>% group_by(applicant_race_and_ethnicity) %>%
  summarise(EthnicityCount = n()) %>%
  arrange(desc(EthnicityCount))

graph_by_enthicity(mortgage_by_race_and_ethnicity)
```

# Graph which applicant races and ethnicities have the largest proportion of loans
# in various stages. These include origination status, denied status, etc.

```{r}
mortgage_status_by_race_and_ethnicity <- hmda_data_pa_df %>% group_by(action_taken_name, applicant_race_and_ethnicity) %>%
  summarise(ActionCount = n()) %>%
  arrange(desc(ActionCount))

mortgage_status_aggregated_by_race_and_ethnicity  = inner_join(mortgage_status_by_race_and_ethnicity, mortgage_by_race_and_ethnicity) %>% mutate(percentage = (ActionCount / EthnicityCount) * 100)

graph_application_race_proportion_of_loans(mortgage_status_aggregated_by_race_and_ethnicity)

```

# Applicant income histograms.

```{r}
hmda_origination_status_df <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1", ]
graph_applicant_income_histogram(hmda_origination_status_df, "Applicant income distribution for originated loans")
```

# Graph median income for originated loans.
```{r}
hmda_origination_status_df <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1", ]

head(hmda_origination_status_df)

hmda_origination_status_df %>% ggplot(aes(as.numeric(hud_median_family_income))) +
geom_histogram(binwidth = 1000,, fill=c("blue")) + labs(x = "Median Income", y = "Applicant Count", title = "Median Income Distribution for Area for Originated Loans") + theme_bw()
```

## Graph loan distribution by county.

```{r}
mortgage_distribution_by_counties <- hmda_data_pa_df %>%
  filter(!is.na(county_name)) %>%
  group_by(county_name) %>%
  summarise(CountLoans = n() ) %>%
  mutate(percentage = ( CountLoans / sum(CountLoans) ) * 100 ) %>%
  mutate(county_name = reorder(county_name, percentage)) %>%
  arrange(desc(percentage)) %>%
  head(20)

graph_distribution_by_county(mortgage_distribution_by_counties)

originated_mortgage_distribution_by_counties <- hmda_origination_status_df %>%
  filter(!is.na(county_name)) %>%
  group_by(county_name) %>%
  summarise(CountLoans = n() ) %>%
  mutate(percentage = ( CountLoans / sum(CountLoans) ) *100 ) %>%
  mutate(county_name = reorder(county_name, percentage)) %>%
  arrange(desc(percentage)) %>%
  head(20)

graph_distribution_by_county(originated_mortgage_distribution_by_counties)

```

## Graph home loan application distribution for the top 4 counties in the above
## chart by applicant_race_1
```{r}

county_names <- c("Allegheny County", "Philadelphia County", "Montgomery County", "Bucks County")

for (county_name in county_names) {
  
  hmda_data_county_df <- hmda_data_pa_df[hmda_data_pa_df$county_name == county_name, ]
  
  mortgage_by_race_county <- hmda_data_county_df %>% group_by(applicant_race_name_1) %>%
    summarise(RaceCount = n()) %>% arrange(desc(RaceCount))
  
  print(graph_mortgage_distribution_by_race1(mortgage_by_race_county))
}

```

# Graph which communities have the largest proportion of loans in various stages.
# for the top 4 counties listed above.
# These include origination status, denied status, etc.

```{r}

for (county_name in county_names) {
  
  hmda_data_county_df <- hmda_data_pa_df[hmda_data_pa_df$county_name == county_name, ]
  
  mortgage_by_race1_county <- hmda_data_county_df %>% group_by(applicant_race_and_ethnicity) %>%
    summarise(RaceCount = n()) %>% arrange(desc(RaceCount))

  mortgage_status_by_race1_by_county <- hmda_data_county_df %>% group_by(action_taken_name, applicant_race_and_ethnicity) %>%
    summarise(ActionCount = n()) %>%
    arrange(desc(ActionCount))
  
  mortgage_status_aggregated_by_race1_by_county  = inner_join(mortgage_status_by_race1_by_county, mortgage_by_race1_county) %>% mutate(percentage = (ActionCount / RaceCount) * 100)
  
  print(graph_application_race_and_ethnicity_proportion_of_loans(mortgage_status_aggregated_by_race1_by_county))
}

```

## Graph income distribution for Whites and African Americans per county for the
## top 4 counties above.
```{r}

for (county_name in county_names) {
  hmda_origination_status_df_by_county_white <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1" & hmda_data_pa_df$county_name == county_name & hmda_data_pa_df$applicant_race_and_ethnicity == "White", ]
  print(graph_applicant_income_histogram(hmda_origination_status_df_by_county_white, "Income distribution for whites"))

  hmda_origination_status_df_by_county_african_american <- hmda_data_pa_df[hmda_data_pa_df$action_taken == "1" & hmda_data_pa_df$county_name == county_name & hmda_data_pa_df$applicant_race_and_ethnicity == "Black or African American", ]
  print(graph_applicant_income_histogram(hmda_origination_status_df_by_county_african_american, "Income distribution for African Americans"))
}

```


## Graph home loan application distribution for the top 4 counties in the above
## chart by applicant_race_and_ethnicity
```{r}

county_names <- c("Allegheny County", "Philadelphia County", "Montgomery County", "Bucks County")

for (county_name in county_names) {
  
  hmda_data_county_df <- hmda_data_pa_df[hmda_data_pa_df$county_name == county_name, ]
  
  mortgage_by_race_county <- hmda_data_county_df %>% group_by(applicant_race_and_ethnicity) %>%
    summarise(RaceCount = n()) %>% arrange(desc(RaceCount))
  
  print(graph_mortgage_distribution_by_race_and_ethnicity(mortgage_by_race_county))
}

```

# Graph which communities have the largest proportion of loans in various stages.
# for the top 4 counties listed above.
# These include origination status, denied status, etc.

```{r}
for (county_name in county_names) {
  
  hmda_data_county_df <- hmda_data_pa_df[hmda_data_pa_df$county_name == county_name, ]
  
  mortgage_by_race1_county <- hmda_data_county_df %>% group_by(applicant_race_and_ethnicity) %>%
    summarise(RaceCount = n()) %>% arrange(desc(RaceCount))

  mortgage_status_by_race1_by_county <- hmda_data_county_df %>% group_by(action_taken_name, applicant_race_and_ethnicity) %>%
    summarise(ActionCount = n()) %>%
    arrange(desc(ActionCount))
  
  mortgage_status_aggregated_by_race1_by_county  = inner_join(mortgage_status_by_race1_by_county, mortgage_by_race1_county) %>% mutate(percentage = (ActionCount / RaceCount) * 100)
  
  print(graph_application_race_and_ethnicity_proportion_of_loans(mortgage_status_aggregated_by_race1_by_county))
}
```
## Visualize missing variables and impute as needed.

```{r}
library(mice)
library(naniar)

# https://cran.r-project.org/web/packages/naniar/vignettes/naniar-visualisation.html
gg_miss_upset(hmda_data_pa_df)

hmda_data_pa_df_imputed <- mice(hmda_data_pa_df, m=1, maxit=2, meth='cart',seed=500)

hmda_data_pa_df_imputed <- mice::complete(hmda_data_pa_df_imputed)
```


## More analysis on the imputed dataset.
```{r}
summary(hmda_data_pa_df_imputed)
gg_miss_upset(hmda_data_pa_df_imputed)
```

# Additional analysis on the hmda dataset. correlation matrix, plots, etc.
```{r}
# https://stackoverflow.com/questions/20637360/convert-all-data-frame-character-columns-to-factors
hmda_data_pa_df$loan_to_income_ratio <- hmda_data_pa_df$loan_amount_000s / hmda_data_pa_df$applicant_income_000s

hmda_data_pa_df[sapply(hmda_data_pa_df, is.character)] <- lapply(hmda_data_pa_df[sapply(hmda_data_pa_df, is.character)], 
                                       as.factor)

hmda_data_pa_df_for_correlation <- as.data.frame(lapply(hmda_data_pa_df, as.integer))

#head(hmda_data_pa_df_for_correlation[, c("applicant_income_000s", "loan_amount_000s")])

head(hmda_data_pa_df_for_correlation)

corr_simple(hmda_data_pa_df_for_correlation)

corrplot(cor(hmda_data_pa_df_for_correlation[, c("applicant_income_000s", "loan_amount_000s")], use = "na.or.complete"))

```

# Additional analysis on the hmda imputed dataset. correlation plots, etc.
```{r}
# hmda_data_pa_df_imputed <- hmda_data_pa_df;
# https://stackoverflow.com/questions/20637360/convert-all-data-frame-character-columns-to-factors
hmda_data_pa_df_imputed$loan_to_income_ratio <- hmda_data_pa_df_imputed$loan_amount_000s / hmda_data_pa_df_imputed$applicant_income_000s

hmda_data_pa_df_imputed[sapply(hmda_data_pa_df_imputed, is.character)] <- lapply(hmda_data_pa_df_imputed[sapply(hmda_data_pa_df_imputed, is.character)], 
                                       as.factor)

hmda_data_pa_df_imputed_for_correlation <- as.data.frame(lapply(hmda_data_pa_df_imputed, as.integer))

head(hmda_data_pa_df_imputed_for_correlation[, c("applicant_income_000s", "loan_amount_000s")])

corr_simple(hmda_data_pa_df_imputed_for_correlation)

corrplot(cor(hmda_data_pa_df_imputed_for_correlation[, c("applicant_income_000s", "loan_amount_000s")], use = "na.or.complete"))

```

# Save the imputed dataframe to file.
```{r}
write.csv(hmda_data_pa_df_imputed, "D:/data/HMDA/2015/hmda_2015_pa_imputed.csv", row.names = FALSE)
```

## Initial Model
## Selecting Columns

```{r}
data_frame_for_model <- function(input_df) {
  # Added based on intuition. Please update as needed.
  # msamd -> Metropolitan Statistical Area Division code.
  # https://cfpb.github.io/api/hmda/fields.html
  initial_columns_for_model <- c("applicant_race_and_ethnicity", "applicant_income_000s", "co_applicant_sex", "county_code", "hoepa_status", "lien_status", "msamd", "owner_occupancy", "preapproval", "property_type", "purchaser_type", "loan_amount_000s", "county_name", "applicant_sex", "hud_median_family_income", "tract_to_msamd_income", "loan_to_income_ratio", "agency_code")

  # https://stackoverflow.com/questions/10085806/extracting-specific-columns-from-a-data-frame
  df_for_model <- input_df %>% select(initial_columns_for_model)
  df_for_model$loan_granted <- input_df$action_taken == 1 | input_df$action_taken == 6
  return (df_for_model)
}
```


# Create dataframe based on selected column list from the source dataframe.

```{r}
hmda_data_pa_df_imputed_for_model <- data_frame_for_model(hmda_data_pa_df_imputed)
```


```{r}
unique(hmda_data_pa_df_imputed$applicant_race_and_ethnicity)
```


```{r}
nrow(hmda_data_pa_df_imputed_for_model)
```
## Training and Testing Set.
We use Stratified Sampling on the predictor whether loan is granted or not.

```{r}
train_test_split <- function(input_df, train_proportion = 0.8) {
  library(caret)
  
  # https://stackoverflow.com/questions/20776887/stratified-splitting-the-data
  train.index <- createDataPartition(input_df$loan_granted, p = train_proportion, list = FALSE)

  train <- input_df[train.index, ]
  test <-  input_df[-train.index, ]
  
  train$loan_granted <- as.factor(train$loan_granted)
  test$loan_granted <- as.factor(test$loan_granted)
  
  val_list <- list(train, test)
  return (val_list)
}
```


```{r}
writeLines("Number of Applicants granted loan")
print(sum(hmda_data_pa_df_imputed_for_model$loan_granted == TRUE))

writeLines("Number of Applicants granted denied")
print(sum(hmda_data_pa_df_imputed_for_model$loan_granted == FALSE))

print(nrow(hmda_data_pa_df_imputed))
```


```{r}
split_dfs <- train_test_split(hmda_data_pa_df_imputed_for_model[, -which(names(hmda_data_pa_df_imputed_for_model) %in% c("co_applicant_race_and_ethnicity", "county_name"))])
train <- split_dfs[[1]]
test <- split_dfs[[2]]

head(train)
```


```{r}
levels(hmda_data_pa_df_imputed_for_model$applicant_race_and_ethnicity)
```

## Race and ethnicity numbers for whites and African Americans
```{r}
sum(train$applicant_race_and_ethnicity == "White")
```


```{r}
```


```{r}
sum(train$applicant_race_and_ethnicity == "Black or African American")
```
### Generating upsampled training dataframe based on race and ethnicity
```{r}
set.seed(1000)

up_train <- upSample(x = train, y = train$applicant_race_and_ethnicity)

sum(as.character(up_train$applicant_race_and_ethnicity) == "White")

sum(as.character(up_train$applicant_race_and_ethnicity) == "Black or African American")
```


```{r}
nrow(up_train)
```


```{r}
sum(test$applicant_race_and_ethnicity == "White")
sum(test$applicant_race_and_ethnicity == "Black or African American")
nrow(test)
```

# Data Models
## Model 1: Without using loan to income ratio
```{r}
# Initial model. Predictors selected are based on intuition.
# Please update this list as needed.

model <- glm(loan_granted~applicant_race_and_ethnicity+applicant_income_000s+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code, family = binomial(link='logit'), data=train)
```

Summarry for the above model
```{r}
summary(model)
```

ANOVA Test
```{r}
anova(model)
```


```{r}
plot(model)
```
## Predictions on this model
```{r}
fitted.probs <- predict(model, test, type = "response")
```
## Confusion Matrix, Classification error and few other statistics

```{r}
prediction_stats <- function(fitted.probs, threshold = 0.5) {
  fitted.results <- ifelse(fitted.probs > 0.5, TRUE, FALSE)
  
  classification_error <- mean(fitted.results != test$loan_granted)
  classification_error
  
  print(paste('Accuracy', 1 - classification_error))
  
  print(confusionMatrix(as.factor(fitted.results), as.factor(test$loan_granted)))
}

```
## Model Performance
```{r}
prediction_stats(fitted.probs)

```
## ROC Curve
```{r}
library(ROCR)
pred <- prediction(fitted.probs, test$loan_granted)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf)
```






## Model 2: Considering loan to income ratio
```{r}
model_loan_income_ratio <- glm(loan_granted~applicant_race_and_ethnicity+applicant_income_000s+loan_to_income_ratio+hoepa_status+lien_status+owner_occupancy+preapproval+property_type+loan_amount_000s+msamd+applicant_sex+county_code+hud_median_family_income+tract_to_msamd_income+agency_code, family = binomial(link='logit'), data=train)
```


```{r}
summary(model_loan_income_ratio)
```


```{r}
anova(model_loan_income_ratio)
```
## Model Plots
```{r}
plot(model_loan_income_ratio)
```
## Model Performance
```{r}
fitted_2.probs <- predict(model_loan_income_ratio, test, type = "response")
prediction_stats(fitted_2.probs)
```
## ROC Curve

```{r}
pred <- prediction(fitted_2.probs, test$loan_granted)
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf)
```
