---
title: "HMDA_2008_IL_EDA - Initial Analysis for 2008 IL Data"
author: "Virat Joshi"
date: "4/5/2020"
output: html_document
---

### Set working directory
```{r}
working_dir = "/Users/vjstark/Downloads/CSP-571/Project"
setwd(working_dir)

data_dir <- "./Data"
```

## Load essential Libraries
```{r}
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(corrplot)
library(data.table)
library(mice)
library(rstudioapi)    
#library(naniar)
```
## Load data file
```{r}
hmda_data_pa <- fread(paste(data_dir, "/hmda_2007_pa_all-records_labels.csv", sep = ""))

```
## Exploratory Data Analysis
```{r}
hmda_pa_df <- as.data.frame(hmda_data_pa)
head(hmda_pa_df)
```

```{r}
colnames(hmda_pa_df)
```

```{r}
summary(hmda_pa_df)
```
```{r}
df_copy<-hmda_pa_df
```

```{r}
dim(hmda_pa_df)
```

```{r}
glimpse(hmda_pa_df)
```

## Handle missing data
```{r}
sapply(hmda_pa_df, function(x) sum(is.na(x)))
```

Check for any empty strings

```{r}
sapply(hmda_pa_df, function(x) sum(x == ""))
```

```{r}
sapply(hmda_pa_df, function(x) sum(x == NULL))
```

## Insights into applicant race and ethnicity columns
```{r}
unique(hmda_pa_df$applicant_race_name_1)
```

```{r}
unique(hmda_pa_df$applicant_ethnicity_name)
```

```{r}
table(hmda_pa_df$applicant_race_name_1)
```

```{r}
hmda_pa_df$applicant_race_and_ethnicity <- NA
hmda_pa_df$co_applicant_race_and_ethnicity <- NA

hmda_pa_df$applicant_race_and_ethnicity <- ifelse(hmda_pa_df$applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_pa_df$applicant_race_name_1)

hmda_pa_df$co_applicant_race_and_ethnicity <- ifelse(hmda_pa_df$co_applicant_ethnicity_name == "Hispanic or Latino",
       "Hispanic or Latino", hmda_pa_df$co_applicant_race_name_1)

head(hmda_pa_df)
```

```{r}
count_racename <- as.data.frame(table(hmda_pa_df$applicant_race_and_ethnicity))
colnames(count_racename)<-c("Race_and_Ethnicity","Count")

p<-ggplot(data=count_racename, aes(x=reorder(Race_and_Ethnicity,-Count), y=Count)) + geom_bar(stat="identity",fill="steelblue")+geom_text(aes(label=Count), size=2.5)
p + coord_flip() +labs(title = "Distribution of applicants by Race and Ethnicity", y="Number of Applicants",x="Race and Ethnicity")

#text(x = p, y = count_racename$Count, label = count_racename$Count, pos = 3, cex = 0.8, col = "red")
```

```{r}
count_racename[order(count_racename$Count),]
```

```{r}
hmda_origination_status_df <- hmda_pa_df[hmda_pa_df$action_taken == "1", ]
head(hmda_origination_status_df)
```

```{r}
ggplot(hmda_origination_status_df, aes(x=applicant_income_000s)) + geom_histogram(binwidth = 10)
```


```{r}
max(hmda_origination_status_df$applicant_income_000s)-min(hmda_origination_status_df$applicant_income_000s)
```

```{r}
hist(hmda_origination_status_df$applicant_income_000s)
```

```{r}
class(hmda_origination_status_df)

sum(is.na(hmda_origination_status_df$applicant_income_000s))/length(hmda_origination_status_df$applicant_income_000s)*100

hmda_origination_status_df$applicant_income_000s[is.na(hmda_origination_status_df$applicant_income_000s)]<-median(hmda_origination_status_df$applicant_income_000s,na.rm = TRUE)


length(hmda_origination_status_df$applicant_income_000s[hmda_origination_status_df$applicant_income_000s>1000])
```

## Income of applicants
```{r}
summary(hmda_pa_df$applicant_income_000s)
```
Here it can be observed that the income of applicants is spreadout a lot. We have applicants with income ranging from 1 to 9999 in thousands. Also, missing values sum upto 62522.

We can replace missing values in applicants income with median.

```{r}
hmda_pa_df$applicant_income_000s[is.na(hmda_pa_df$applicant_income_000s)]<-median(hmda_pa_df$applicant_income_000s,na.rm = TRUE)
```

Find the income distribution based on race and ethnicity
```{r}
p <- ggplot(group_by(hmda_pa_df,applicant_race_and_ethnicity), aes(x=applicant_race_and_ethnicity, y=applicant_income_000s)) + 
  geom_boxplot()
p +coord_flip()
```

In the boxplot, the category where people didn't provide the race and ethnicity data has the second most extreme values. 
### These are probably outliers and we need to deal with them

```{r}
unique(hmda_pa_df$action_taken_name)
```

Here we are trying to find the distribution of loans that are denied over race and ethnicity.
## Create a new column where we classify the action taken as loan is denied or not.

```{r}
hmda_pa_df$denied
hmda_pa_df$denied<-NA

hmda_pa_df$denied[hmda_pa_df$action_taken_name == "Application denied by financial institution" | hmda_pa_df$action_taken_name == "Preapproval request denied by financial institution"]<- "Denied"

hmda_pa_df$denied[is.na(hmda_pa_df$denied)]<-"Approved"
unique(hmda_pa_df$denied)

application_denied<-sum(hmda_pa_df$denied == "Denied")
application_originated<-sum(hmda_pa_df$denied == "Approved")
application_denied/length(hmda_pa_df$action_taken)

ggplot(hmda_pa_df,aes(hmda_pa_df$applicant_race_and_ethnicity, ..count..)) +geom_bar(aes(fill= hmda_pa_df$denied))

class(hmda_pa_df$denied)
p<-ggplot(hmda_pa_df , aes (x = applicant_race_name_1 , fill = denied,)) + geom_bar(position = "dodge")
p + coord_flip()
hmda_pa_df$denied
```