---
title: "Homeownership Rate in USA"
output: html_notebook
author: "Omkar Pawar"

---

## Import Data from Census Homeownership Rate according to race.

```{r}
library("readxl")
library(ggplot2)
df <- read_excel("HomeOwnershipByRace.xlsx")
df

df<-na.omit(df)

new_df <- as.matrix(df[-1])
row.names(new_df)<-df$...1
new_df <- as.data.frame(new_df)

new_df
```
Now we have data loaded. From the column names we can see that the first column has the race and remaining columns are per year home ownership rates.
```{r}
df1<-as.data.frame(t(new_df))
```

```{r}
df1 <- cbind(Year = rownames(df1), df1)
rownames(df1) <- 1:nrow(df1)
names(df1)<-make.names(names(df1),unique = TRUE)
df1
colnames(df1)<-c("Year","Overall_US", "All_Whites","Non_Hispanic_Whites","African_American","Minorities",
  "American.Indian.or.Alaskan.Native.alone","Asian.or.Native.Hawaiian.Pacific.Islander.alone","Hispanic.or.Latino","Non.Hispanic")
```
## US Homeownership Rate and Minorities Homeownership Rate
```{r}
ggplot(data = df1 , aes(x = Year , group = 1) ) + 
  geom_line(aes(y = Overall_US , colour = "Overall_US"),linetype = "dashed") + 
  geom_line(aes(y = Minorities , colour = "Minorities"),size = 1)+
  geom_rect(data=rect, aes(xmin=xmin, xmax=xmax,ymin = ymin,ymax = ymax),
              fill="grey10",
              alpha=0.25,
              inherit.aes = FALSE)+
  scale_colour_manual("", 
                      breaks = c("Overall_US","Minorities"),
                      values = c("tomato2", "#009E73"))+
  scale_y_continuous("HomeOwnership Rate", limits = c(50,70))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "Homeownership Rate Comparision between Overall Population and Minorities")
ggsave("/Users/omkarpawar/Desktop/HomeOwnership_Minorirtes.png",width = 14.5,height = 7)
```

## Homeownership Rate Comparision between Whites and Black population
```{r}

rect <- data.frame(xmin=as.factor(2006), xmax=as.factor(2008),ymin = 50,ymax = 70)

ggplot(data = df1 , aes(x = Year , group = 1) ) + 
  geom_line(aes(y = Overall_US , colour = "Overall US"),linetype = "dashed") + 
  geom_line(aes(y = African_American , colour = "African American"),size = 1)+
  geom_line(aes(y = Non_Hispanic_Whites , colour = "Whites"),size = 1)+
  geom_rect(data=rect, aes(xmin=xmin, xmax=xmax,ymin = ymin,ymax = ymax),
              fill="grey10",
              alpha=0.25,
              inherit.aes = FALSE)+
  scale_colour_manual("", 
                      breaks = c("Overall US",  "African American","Whites"),
                      values = c("tomato2", "#009E73","slateblue"))+
  scale_y_continuous( limits = c(40,85))+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "Homeownership Rate Comparision between Whites and African Americans",y = "Home-ownership Rate(%)",x = "Timeline",fill = "Race and Ethnicity")+theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("/Users/omkarpawar/Desktop/HomeOwnership.png",width = 14.5,height = 7)
```


```{r}
ggplot(data = df1 , aes(x = Year , group = 1) ) + 
  geom_line(aes(y = Overall_US , colour = "Overall_US")) + 
  
  geom_line(aes(y = Non_Hispanic_Whites , colour = "Non_Hispanic_Whites"))+
  geom_line(aes(y = African_American , colour = "African_American"))+
  
  geom_line(aes(y = American.Indian.or.Alaskan.Native.alone , colour = "American.Indian.or.Alaskan.Native.alone"))+
  geom_line(aes(y = Asian.or.Native.Hawaiian.Pacific.Islander.alone , colour = "Asian.or.Native.Hawaiian.Pacific.Islander.alone"))+
  geom_line(aes(y = Hispanic.or.Latino , colour = "Hispanic.or.Latino"))+
  
  scale_colour_manual("", 
                      breaks = c("Overall_US", "Non_Hispanic_Whites" , "African_American","American.Indian.or.Alaskan.Native.alone","Asian.or.Native.Hawaiian.Pacific.Islander.alone","Hispanic.or.Latino"),
                      values = c("#999999",  "#56B4E9",  "purple", "#0072B2", "#D55E00", "#CC79A7"))+
  scale_y_continuous("HomeOwnership Rate", limits = c(40,80))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
```

## All Possible Race and Ethnicity Values
```{r}
ggplot(data = df1 , aes(x = Year , group = 1) ) + 
  geom_line(aes(y = Overall_US , colour = "Overall_US")) + 
  geom_line(aes(y = All_Whites , colour = "All_Whites"))+
  geom_line(aes(y = Non_Hispanic_Whites , colour = "Non_Hispanic_Whites"))+
  geom_line(aes(y = African_American , colour = "African_American"))+
  geom_line(aes(y = Minorities , colour = "Minorities"))+
  geom_line(aes(y = American.Indian.or.Alaskan.Native.alone , colour = "American.Indian.or.Alaskan.Native.alone"))+
  geom_line(aes(y = Asian.or.Native.Hawaiian.Pacific.Islander.alone , colour = "Asian.or.Native.Hawaiian.Pacific.Islander.alone"))+
  geom_line(aes(y = Hispanic.or.Latino , colour = "Hispanic.or.Latino"))+
  geom_line(aes(y = Non.Hispanic , colour = "Non.Hispanic"))+
  scale_colour_manual("", 
                      breaks = c("Overall_US", "All_Whites","Non_Hispanic_Whites" , "African_American","Minorities","American.Indian.or.Alaskan.Native.alone","Asian.or.Native.Hawaiian.Pacific.Islander.alone","Hispanic.or.Latino","Non.Hispanic"),
                      values = c("#999999", "#E69F00", "#56B4E9", "#009E73", "purple", "#0072B2", "#D55E00", "#CC79A7","#293352"))+
  scale_y_continuous("HomeOwnership Rate", limits = c(40,80)) +
theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "All Possible Race and Ethnicity Values")

```



