---
title: "IL 2017 Logistic Regression"
author: "Omkar Pawar"
output: html_notebook
---

## We tried the following glm models. The predictors chosen were picked
## from our EDA analysis and feature engineering work and some background
## knowledge.
# Model 1. Raw loan and applicant income.
# Model 2. Log of loan and applicant income.
# Model 3. Log of loan amount and loan to income ratio.
# Model 4. Cross validation model with same predictors as Model 3.

## Results.
# Model 2, Model 3 and Model 4 performed similarly. The criterion was AUC 
# curve, Mcfadden R square value and confusion matrix results on test data.

# Model 3 was chosen as banks do use the loan to income ratio as one of their
# criterion. 

## Key observation
# The AUC curve for our models ranges between 74 to 77%. The AUC curve can
# be thought of as a classification-threshold-invariant. It measures the #
# quality of the model's predictions irrespective of the threshold value
# https://developers.google.com/machine-learning/crash-course/classification/roc-and-auc.
# We feel that this is because our models miss crucial data like debt to income
# ratio, credit scores, etc which banks use as additional predictors.


```{r}
library(sys)

working_directory <- getwd()

setwd(dirname(dirname(working_directory)))

# The models are saved in this directory.
model_path <- paste(dirname(rstudioapi::getActiveDocumentContext()$path), "models", sep = "/")
print(model_path)

library(corrplot)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(data.table)
library(mice)
library(rstudioapi)    
library(naniar)
library(caret)
library(pROC)
library(ROCR)
library(lmtest)

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/utils.r", sep="/"))

source(paste(dirname(dirname(dirname(rstudioapi::getActiveDocumentContext()$path))), "utils/model_utils.r", sep="/"))
```

```{r}
hmda_data_df_ <- fread("/Users/omkarpawar/Desktop/imputed_il_2017.csv")
hmda_data_df <- as.data.frame(hmda_data_df_)
```


# Create dataframe based on selected column list from the source dataframe.
```{r}

df_for_model <- hmda_data_frame_for_model(hmda_data_df)
df_for_model <- process_model_df_columns(df_for_model)

```

```{r}
head(df_for_model, 20)

nrow(df_for_model)
```
```{r}
set.seed(400)

split_dfs <- train_test_split(df_for_model$loan_granted, df_for_model)

train <- split_dfs[[1]]
test <- split_dfs[[2]]
```

# Null model
```{r}
null_model <- build_and_evaluate_hmda_null_model("loan_granted~1", train, test)

```

# Model with applicant income and loan amount.

```{r}
model_2017_income_loan_amount <-
build_glm_model("loan_granted~applicant_race_and_ethnicity+applicant_income_000s+loan_amount_000s+preapproval+co_applicant_present+county_code+minority_population+loan_purpose+tract_to_msamd_income+applicant_sex+co_applicant_sex+property_type+owner_occupancy+agency_code+respondent_id+rate_spread", train)

model_summary_and_plot(model_2017_income_loan_amount, train)
mcfadden_r_square(model_2017_income_loan_amount, null_model)

```
# Model with logs of applicant income and loan amount.
```{r}
model_2017_log_income_loan_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+log_applicant_income_000s+log_loan_amount_000s+preapproval+co_applicant_present+county_code+tract_to_msamd_income+loan_purpose+owner_occupancy+respondent_id+minority_population+rate_spread+agency_code+applicant_sex+co_applicant_sex+property_type", train)

model_summary_and_plot(model_2017_log_income_loan_amount, train)
mcfadden_r_square(model_2017_log_income_loan_amount, null_model)

```
# Model with logs of applicant income and loan amount and loan to income ratio

```{r}
model_2017_ratio_of_loan_to_income_with_log_of_income_and_amount <- build_glm_model("loan_granted~applicant_race_and_ethnicity+owner_occupancy+preapproval+log_loan_amount_000s+applicant_sex+county_code+tract_to_msamd_income+co_applicant_present+agency_code+minority_population+loan_to_income_ratio+co_applicant_sex+loan_purpose+rate_spread+property_type+log_applicant_income_000s", train)

model_summary_and_plot(model_2017_ratio_of_loan_to_income_with_log_of_income_and_amount, train)
mcfadden_r_square(model_2017_ratio_of_loan_to_income_with_log_of_income_and_amount, null_model)

```
# Cross validation
```{r}
model_2017_cross_val_loan_income_ratio <- cross_validation("loan_granted~applicant_race_and_ethnicity+loan_to_income_ratio+co_applicant_present+owner_occupancy+preapproval+property_type+respondent_id+log_loan_amount_000s+county_code+loan_purpose+tract_to_msamd_income+minority_population+respondent_id+agency_code+rate_spread+log_applicant_income_000s", train)

model_summary_and_plot(model_2017_cross_val_loan_income_ratio$finalModel, train)

```

# Model performance.

```{r}
library(ROCR)
library(pROC)
model1_fitted.probs <- predict(model_2017_income_loan_amount, test, type = "response")
model2_fitted.probs <- predict(model_2017_log_income_loan_amount, test, type = "response")
model3_fitted.probs <- predict(model_2017_ratio_of_loan_to_income_with_log_of_income_and_amount, test, type = "response")

writeLines("")
writeLines("Model 1 performance")
print_model_performance(model1_fitted.probs, test, positive = "yes", negative = "no", threshold = 0.74)

writeLines("")
writeLines("Model 2 performance")
print_model_performance(model2_fitted.probs, test, positive = "yes", negative = "no", threshold = 0.74)

writeLines("")
writeLines("Model 3 performance")
print_model_performance(model3_fitted.probs, test, positive = "yes", negative = "no", threshold = 0.75)

```
# Cross validation performance
```{r}
library(pROC)

model4_fitted.probs <- predict(model_2017_cross_val_loan_income_ratio, test, type = "prob")
writeLines("")
writeLines("Cross validation model performance")

writeLines("")
writeLines("Model 3 performance")
print_model_performance(model4_fitted.probs[, 2], test, positive = "yes", negative = "no", threshold = 0.6)
```
